Toward Stable Replication of Genomic Information in Pools of RNA Molecules

Version 1

Abstract

The transition from prebiotic chemistry to living systems requires the emergence of a scheme for enzyme-free genetic replication. Here, we analyze a recently proposed prebiotic replication scenario, the so-called Virtual Circular Genome (VCG) [Zhou et al., RNA 27, 1-11 (2021)]: Replication takes place in a pool of oligomers, where each oligomer contains a subsequence of a circular genome, such that the oligomers encode the full genome collectively. While the sequence of the circular genome may be reconstructed based on long oligomers, short oligomers merely act as replication feedstock. We observe a competition between the predominantly error-free ligation of a short oligomer to a long oligomer and the predominantly erroneous ligation of two long oligomers. Increasing the length of long oligomers and reducing their concentration decreases the fraction of erroneous ligations, enabling high-fidelity replication in the VCG. Alternatively, the problem of erroneous products can be mitigated if only monomers are activated, such that each ligation involves at least one monomer. Surprisingly, in such systems, shorter oligomers are extended by monomers more quickly than long oligomers, a phenomenon which has already been observed experimentally [Ding et al., JACS 145, 7504-7515 (2023)]. Our work provides a theoretical explanation for this behavior, and predicts its dependence on system parameters such as the concentration of long oligomers. Taken together, the VCG constitutes a promising scenario of prebiotic information replication: It could mitigate challenges of in non-enzymatic copying via template-directed polymerization, such as short lengths of copied products and high error rates.

Introduction

In order to delineate possible pathways towards the emergence of life, it is necessary to understand how a chemical reaction network capable of storing and replicating genetic information might arise from prebiotic chemistry. RNA is commonly assumed to play a central role on this path, as it can store information in its sequence and catalyze its own replication [1–3]. While ribozymes capable of replicating strands of their own length have been demonstrated in the laboratory [4], it remains elusive how enzyme-free self-replication might have worked before the emergence of such complex ribozymes.

One possible mechanism is template-directed primer extension [5–11]: In this process, a primer hybridizes to a template and is extended by short oligonucleotides, thereby forming a (complementary) copy of the template strand. Considerable progress has been made in optimizing template-directed primer extension, but challenges remain: (i) The produced copies are likely to be incomplete. So far, at most 12 nt have been successfully added to an existing primer [10]. Moreover, as the pool of primer strands needs to emerge via random polymerization, the primer is likely to hybridize to the template at various positions, and not only to its 3’-end, leaving part of the 3’-end of the template uncopied [12]. (ii) Errors in enzyme-free copying are frequent due to the limited thermodynamic discrimination between correct Watson-Crick pairing and mismatches [13–16] The error probability constrains the length of the genome that can be reliably replicated.

The issue of insufficient thermodynamic discrimination can, in principle, be mitigated by making use of kinetic stalling after the incorporation of a mismatch [15]. By introducing a competition between the reduced polymerization rate and a characteristic timescale of the non-equilibrium environment, it is possible to filter correct sequences from incorrect ones [16]. To address the challenge of incomplete copies, Zhou et al. propose an new scenario of replication, the so-called Virtual Circular Genome (VCG) [17]. In this scenario, genetic information is stored in a pool of oligomers that are shorter than the circular genome they collectively encode: Each oligomer bears a subsequence of the circular genome, such that the collection of all oligomers encodes the full circular genome virtually. Within the pool, each oligomer can act as template or primer [17]. The oligomers hybridize to each other and form complexes that allow for templated ligation of two oligomers, or for the extension of an oligomer by a monomer. Because the sequences of the ligated strands and the template are part of the genome sequence, most of the products should also retain the sequence of the genome. That way, long oligomers encoding the circular genome are produced at the expense of short oligomers [17]. The long strands, in turn, can assemble into catalytically active ribozymes. With a continuous influx of short oligomers, the VCG might allow for continuous replication of the virtually encoded circular genome. Importantly, replication in the VCG is expected to avoid the issue of incomplete copies. Since the genome is circular, it does not matter which part of the genome an oligomer encodes, as long as the sequence is compatible with the genome sequence. An additional feature of the VCG scenario is that replication should be achievable without the need of adding many nucleotides to a primer: Provided the concentration of oligomers decreases exponentially with their length, the concentration of each oligomer in the pool can be doubled by extending each oligomer only by a few nucleotides [17]. The extension of an oligomer by a few nucleotides in a VCG pool has already been demonstrated experimentally [18].

A recent computational study points out that the VCG scenario is prone to loss of genetic information via ”sequence scrambling” [19]: If the genome contains identical sequence motifs at multiple different loci, replication in the VCG will mix the sequences of these loci, thus destroying the initially defined genome. More generally, it is currently unclear, which conditions could prevent genome instability of VCGs, such that their genetic information is retained. Length distribution, sequence composition, oligonucleotide concentration and environmental conditions, such as temperature, all affect the stability of complexes and thus the replication dynamics of the VCG pool. Here, we characterize the replication fidelity and yield of VCG pools using a kinetic model, which explicitly incorporates association and dissociation of RNA strands as well as templated ligation. We study a broad spectrum of prebiotically plausible and experimentally accessible oligomer pools, from pools containing only monomers and long oligomers of a single length to pools including a range of long oligomers with uniform or exponential concentration profile. The length of the included oligomers as well as their concentration are free parameters of our model.

We find that, regardless of the details of the pool, there is a competition between three types of template-directed ligation reactions: (i) Ligations of two short oligomers, e.g., monomers, forming oligomers that are too short to specify a unique locus on the genome. (ii) Ligations of a short oligomer to a long oligomer, which form longer oligomers with sequences that are typically compatible to the genome. (iii) Ligations of two long oligomers, which often create oligomers that are incompatible with the genome sequence. Therefore, the fidelity is primarily determined by the competition between the error-free extension of a long oligomer and the erroneous ligation of two long oligomers. The likeli-hood of the latter reaction can be minimized by lowering the relative concentration of long oligomers, even though this increases the fraction of short oligomers that are used up for ligations of the first type. Consequently, the fidelity can be improved at the expense of reducing the yield. The efficiency, meaning the yield attainable at a fixed high fidelity, thus depends on the length distribution of the oligomers in the pool.

Alternatively, the issue of erroneous ligations is solved if only monomers are activated, such that each ligation incorporates only one monomer at a time, as in the experimental study by Ding et al. [18]. In this case, the VCG concentration can be chosen arbitrarily large without compromising fidelity. Interestingly, our model predicts an unexpected feature: In the limit of high VCG concentrations, the rate of primer extension is higher for shorter oligomers than for long oligomers, even though, intuitively, complexes containing longer oligomers are expected to be more stable and thus more productive. The same behavior was indeed observed experimentally [18]. We provide an explanation for this feature, and discuss its dependence on system parameters such as the length and the concentration of long oligomers in the pool.

Model

The VCG is a replication scenario where a circular genome is stored in a pool of oligomers. The oligomers are shorter than the genome that they encode. Each oligomer bears a subsequence of the circular genome, such that, collectively, the oligomers encode the full genome (Fig. 1A). To set up our model for the dynamics of VCG pools, we specify (i) our choice of circular genomes, (ii) how to map the genomes to pools of oligomers, and (iii) which chemical reactions govern the dynamics.

Model. A In the Virtual Circular Genome (VCG) model, a circular genome (depicted in green) as well as its sequence complement are encoded in a pool of oligomers (depicted in blue and orange). Collectively, the pool of oligomers encodes the whole sequence of the circular genome. Depending on their length, two types of oligomers can be distinguished: Feedstock oligomers (depicted in blue) are too short to specify a unique locus on the genome, while long VCG oligomers (depicted in orange) do. B The length-distribution of oligomers included in the VCG pool is assumed to be exponential. The concentration of feedstock oligomers and VCG oligomers as well as their respective length scales of exponential decay and can be varied independently. The set of included oligomer lengths can be restricted via and . C The hybridization energy of complexes is computed using a simplified nearest-neighbor model: Each full block comprised of two base pairs (depicted in pink) contributes γ, while dangling end blocks (depicted in blue) contribute γ/2. D Oligomers form complexes via hybridization reactions, or dehybridize from an existing complex. The ratio of hybridization and dehybridization rate is governed by the hybridization energy. If two oligomers are adjacent to each other in a complex, they can undergo templated ligation. E Based on the length of the reacting oligomers, we distinguish three types of templated ligation: Ligation of two feedstock oligomers (F+F), ligation of feedstock oligomer to VCG oligomer (F+V) and ligation of two VCG oligomers (V+V).

Circular Genomes

For a given genome length, LG, there are distinct circular genomes (the factor 1/2LG accounts for the freedom to select the starting position and to choose the Watson or the Crick strand as reference sequence). A key property of the genome is its minimal unique motif length, LS, defined as the shortest length such that all possible subsequences of length L ≥ LS appear at most once in the genome. In other words, all subsequences of length L ≥ LS specify a unique locus on the genome. In principle, each circular genome has another length scale, corresponding to the maximal motif length, up to which all possible sequence motifs are contained in the genome. Here, for convenience, we restrict ourselves to genomes in which this length-scale is LS − 1, such that all possible subsequences shorter than LS are contained in the sequence of the genome. Additionally, we choose only unbiased genomes in which all possible subsequences of length L < LS are contained at equal frequency.

Construction of VCG Pools

To specify a VCG pool encoding a genomic sequence (chosen as described above), one must select the subsequences to be included in the pool, as well as their concentrations. We consider unbiased pools, where the concentration of subsequences, c(L), depends only on their length, L, i.e., all subsequences of a given length are included at equal concentration. We refer to the length-dependent concentration profile as the length distribution of the pool. Depending on their length, oligomers fall into two categories (Fig. 1B): (i) short feedstock oligomers and (ii) long VCG oligomers. Feedstock oligomers are oligomers that are shorter than the unique subsequence length LS. Since their sequence appears multiple times on the genome, they do not encode a specific position on the genome. Thus, they serve as feedstock for the elongation of VCG oligomers rather than as information storage. Conversely, the VCG oligomers, which are longer than the unique subsequence length LS, have a unique locus on the circular genome. Collectively, the VCG oligomers enable the reconstruction of the full genome.

The full length distribution, c(L), can be decomposed into the contributions of feedstock oligomers and VCG oligomers, We assume that both types of oligomers have an exponential length distribution. In our model, the concentration of VCG oligomers can be varied independently of the concentration of feedstock oligomers, and the length scales for the exponential decay ( vs.) may differ between VCG oligomers and feedstock oligomers. Additionally, we can restrict the set of oligomer lengths included in the pool by setting minimal and maximal lengths for feedstock oligomers and for VCG oligomers individually, For any other oligomer length, the concentrations equal zero. This parametrization includes uniform length distributions as a special case (κF = 0 and κV = 0), and also allows for concentration profiles that are peaked. Peaked length distributions can emerge from the interplay of templated ligation, (de)hybridization and outflux in open systems [20]. We define the total concentration of feedstock oligomers, , as well as the total concentration of VCG oligomers, . Their ratio will turn out to be an important determinant of the VCG dynamics.

(De)hybridization Kinetics

Oligomers can hybridize to each other to form double-stranded complexes, or dehybridize from an existing complex. For simplicity, we do not include self-folding within a strand, which is a reasonable assumption for short oligomers. The stability of a complex is determined by its hybridization energy, with lower hybridization energy indicating greater stability. We use a simplified nearest-neighbor energy model to compute the hybridization energy [20–22]: The total energy equals the sum of the energy contributions of all nearest-neighbor blocks in a given complex (Fig. 1C). The energy contribution associated with a block of two Watson-Crick base pairs (matches) is denoted γ < 0, and dangling end blocks involving one Watson-Crick pair and one free base contribute γ/2. Nearest-neighbor blocks involving mismatches increase the hybridization energy by γMM > 0 per block, thus reducing the stability of the complex. The rate constants of hybridization and dehybridization are related via where c° = 1 M is the standard concentration, and ΔG is the free energy of hybridization. The association rate constant kon is proportional to the encounter rate constant, kenc = 1/(c°t0). The encounter timescale t0 serves as the elementary time unit of the kinetic model, with all reaction timescales measured relative to it.

Templated Ligation

Two oligomers A and B that are hybridized adjacently to each other on a third oligomer C can produce a new oligomer A-B via templated ligation (Fig. 1D). Depending on the length of A and B, we distinguish three types of ligation reactions (Fig. 1E): (i) F+F ligations, in which two feedstock oligomers ligate, (ii) F+V ligations, where a VCG oligomer is extended by a feedstock oligomer, and (iii) V+V ligations involving two VCG oligomers. The formation of a covalent bond via templated ligation is not spontaneous, but requires the presence of an activation reaction. Usually, these reactions add a leaving group to the 5’-end of the nucleotide which is cleaved during bond formation [10, 23]. We assume that the concentration of activating agent is sufficiently high for the activation to be far quicker than the formation of the covalent bond, such that activation and covalent bond formation can be treated as a single effective reaction. We consider pools in which all oligomers are activated, as well as pools in which only monomers are activated.

Observables

Essentially, templated ligation in the pool produces VCG oligomers at the expense of feed-stock oligomers. To characterize these dynamics, we distinguish between different contributions to the production flux. While the product of an F+V ligation is always a VCG oligomer, F+F ligations can produce feed-stock oligomers or VCG oligomers. In both cases, a produced VCG oligomer can be correct (compatible with the genome) or incorrect (incompatible). V+V ligations always produce VCG oligomers that can be correct or incorrect. We quantify these processes by measuring extension fluxes in units of nucleotides ligated to an existing strand (counting the length of the shorter ligated strand as the number of incorporated bases). In particular, we define the fidelity f as the extension flux resulting in correct VCG oligomers relative to the flux resulting in any VCG oligomer, In addition, we introduce the yield y as the proportion of total extension flux that produces VCG oligomers, Efficient replication of the VCG requires both, high fidelity and high yield. Hence, we introduce the efficiency of replication η as the product of fidelity and yield, In order to discern the contributions of different types of templated ligations (F+F, F+V, V+V) to fidelity, yield, and efficiency, we define the ligation share s of a ligation type,

Results

Replication efficiency reaches a maximum at intermediate concentrations of VCG oligomers

To study the dynamics of VCG pools, we start by picking a exemplary genome of length LG = 16 nt,

5’-AAAGAGGACACGGCAU-3’

3’-UUUCUCCUGUGCCGUA-5’.

In this genome, the minimal unique subsequence length is LS = 3 nt. Consequently, monomers as well as dimers are part of the feedstock and VCG oligomers are at least 3 nt long. We first consider the simplest oligomer pool possible, which includes only monomers and VCG oligomers of a single length LV (Fig. 2A).

Replication performance of VCG pools containing VCG oligomers of a single length (single-length VCG pools). A The pool contains a fixed concentration of monomers, mM, as well as VCG oligomers of a single length, LV, at variable concentration (the VCG oligomers cover all possible subsequences of the genome and its complement at equal concentration). B The yield increases as a function of , because dimerizations become increasingly unlikely for high VCG concentrations. C The ligation share of different ligation types depends on the total VCG concentration: In the low concentration limit, dimerization (F+F) dominates; for intermediate concentrations, F+V ligations reach their maximum, while, for high concentrations, a substantial fraction of reactions are V+V ligations. The panel depicts the behavior for LV = 6 nt. D Replication efficiency is limited by the small yield for small . In the limit of high , replication efficiency decreases due to the growing number of error-prone V+V ligations. Maximal replication efficiency is reached at intermediate VCG concentration. E V+V ligations are prone to the formation of incorrect products due to the short overlap between educt strand and template. In general, the probability of correct product formation, pcorr, depends on the choice of circular genome and as well as its mapping to the VCG pool. The probabilities listed here refer to a VCG pool with LG = 16 nt and LS = 3 nt. F The optimal equilibrium concentration ratio of free VCG strands to free feedstock strands, which maximizes replication efficiency, decays as a function of length (continuous line). The analytical scaling law (Eq. (4), dashed line) captures this behavior. The window of close-to-optimal replication, within which efficiency deviates no more than 1% from its optimum (shaded areas), increases with LV, facilitating reliable replication without fine-tuning to match the optimal concentration ratio. G Maximal replication efficiency, which is attained at the optimal VCG concentration depicted in panel E, increases as a function of LV and approaches a plateau of 100%. For high efficiency, Eq. (5) provides a good approximation of the length-dependence of ηmax (dashed lines). The oligomer length at which replication efficiency equals 95% is determined using Eq. (5) (vertical dotted lines). H By construction, the unique subsequence length, LS, increases logarithmically as a function of the length of the genome, LG. The length of VCG oligomers, LV, at which the optimal replication efficiency reaches 95% (computed via Eq. 5) exhibits the same logarithmic dependence on LG.

These initial VCG pools are evolved in time using a stochastic simulation based on the Gillespie algorithm [20–22]. Since the Gillespie algorithm operates on the level of counts of molecules instead of concentrations, we must assign a volume to each system. We choose volumes in the range 1 μm3 to 10 000 μm3 (Supplementary Material Section S-I). In addition to the volume, the reaction rate constants need to be chosen appropriately: (i) Experimentally determined association rate constants are typically around 106 M−1 s−1 [24–26], which corresponds to an association timescale of t0 ≈ 1 μs. (ii) The timescale of dehybridization is computed via Eq. (3) using the energy contribution γ = −2.5 kBT for a matching nearest-neighbor block and γMM = 25.0 kBT in case of mismatches. The high energy penalty of nearest-neighbor blocks involving mismatches, γMM, is chosen to suppress the formation of mismatches, while the value of γ roughly matches the average energy of all matching nearest-neighbor blocks given by the Turner energy model of RNA hybridization [27]. (iii) For templated ligation, we select a reaction rate constant of . This choice of klig is consistent with ligation rates observed in enzyme-free template-directed primer extension experiments, which range from around 10−6 s−1 [9, 10] to roughly 10−4 s−1 [23], depending on the underlying activation chemistry.

Based on the ligation events observed in the time evolution, we compute the observables introduced above (Model section). Due to the small ligation rate constant, the time evolution can only be simulated for a few ligation time units. Consequently, ligation events are scarce, which leads to high variances in the computed observables. We mitigate this issue by calculating the observables based on the concentration of complexes that are in a productive configuration, even if they do not undergo templated ligation within the time window of the simulation (Supplementary Material Section S-I).

From the ligation events, we obtain the total production flux of VCG oligomers and feedstock oligomers. Our observable y (yield) quantifies the fraction of this total flux directed to VCG oligomers. Fig. 2B shows how the yield depends on the concentration of VCG oligomers, , at a fixed total monomer concentration, . Here, the data points (with error bars) represent the simulation results with the different colors corresponding to different choices of initial VCG oligomer length, LV. We observe that the yield increases monotonically with the concentration of VCG oligomers, and approaches 100% for high . The concentration at which the pool reaches a yield of 50% depends on the oligomer length LV: Shorter VCG oligomers require higher concentrations for high yield. The concentration dependence of the yield can be rationalized by the types of templated ligations that are involved. For low VCG concentration, most templated ligation reactions are dimerizations (1+1) with the VCG oligomers merely acting as templates (Fig. 2C). As dimers are shorter than the unique subsequence length LS, their formation does not contribute to the yield, which explains the low yield in the limit of small . Conversely, for high VCG concentration, most of the templated ligations are F+V or V+V ligations, which produce oligomers of length L ≥ LS, implying high yield.

However, Fig. 2C also shows that the relative contribution of V+V ligation increases with increasing , with a large fraction of them producing incorrect products. This reduction in fidelity, f, causes a trade-off between fidelity and yield, which is visible as a peak at intermediate VCG concentrations in the replication efficiency, η = f · y (Fig. 2D). Using hexamers as an example, Fig. 2E illustrates why V+V ligations are prone to forming incorrect oligomers: In order to ensure that two oligomers only ligate if their sequences are adjacent to each other in the true circular genome, both oligomers need to share an overlap of at least LS nucleotides with the template oligomer. Otherwise, the product oligomers might form sequences incompatible with the true genome. The probability of forming incorrect products is a consequence of the combinatorics of possible subsequences in the VCG. Specifically, for a genome with LG = 16 nt, there are 32 different hexamers. For example, if the left educt hexamer only has one nucleotide of overlap with the template, there are 1/4 · 32 = 8 possible educt oligomers. However, only one out of those 8 hexamers is the correct partner for the right educt hexamer, implying an error probability of 1 − 1/8 = 7/8 (first example in Fig. 2E).

Characterizing replication efficiency using simulation data is computationally expensive. Depending on the parameters, up to a hundred simulations, each taking a few days, are needed to obtain a single data point in Fig. 2B-D to sufficient accuracy. To enable rapid computation of the observables of interest over a wide range of parameter values, we introduce an additional approximate approach that relies on (i) the separation of time-scales between (de)hybridization and templated ligation, and (ii) on a coarse-grained representation of the oligomers in the VCG pool. The curves shown together with the simulation data in Fig. 2 are computed using this “adiabatic approach”. The approach is detailed in Supplementary Material Sections S-II and S-III, and only outlined briefly in the following: Due to the slow rate of ligation, the (de)hybridization equilibrium is typically reached long before the first ligation event happens. To determine the hybridization equilibrium, we assume oligomers of the same length to have identical concentration, regardless of their sequence. The concentration of complexes comprised of multiple oligomers is calculated using the law of mass action: It equals the product of the individual oligomer concentrations divided by the respective dissociation constant Kd, which in turn depends on the hybridization energy, ΔG. Depending on the length of the hybridization region, one or multiple hybridization partners may be possible based on the oligomer sequences present in the pool. This is accounted for by weighting the complex concentrations by combinatorial multiplicities. For each oligomer length, we impose a mass conservation constraint ensuring that the equilibrium concentration of free oligomers and the concentration of oligomers bound in complexes adds up to the total concentration in the pool. Solving the law of mass action together with the mass conservation equations numerically yields the equilibrium concentrations of free VCG and feedstock strands, and , for any given total concentration, and . Knowing the equilibrium concentration of free strands allows us to compute concentrations of productive complexes and thus observables characterizing templated ligation.

The results of the adiabatic approach agree well with the data obtained from the simulations (Fig. 2B-D). For instance, the adiabatic approach predicts the same non-monotoneous concentration dependence of replication efficiency as the simulations: VCG pools containing longer oligomers achieve higher maxima in replication efficiency at lower VCG concentration. While the available simulation data only allows for a qualitative description of this trend, the adiabatic approach enables a quantitative analysis. Fig. 2F (solid line) shows the equilibrium concentration ratio at which replication efficiency is maximal, as a function of the VCG oligomer length, LV. As expected from the qualitative trend, pools containing longer oligomers reach their maximum for lower concentration of VCG oligomers. The shaded area indicates the range of VCG concentrations within which the pool’s efficiency deviates by no more than one percent from its optimum. We observe that this range of close-to-optimal VCG concentrations increases with LV. Thus, pools containing longer oligomers require less fine-tuning of the VCG concentration for replication with high efficiency.

In addition to the numerical results, we utilize the adiabatic approach to study the optimal equilibrium VCG concentration analytically (Supplementary Material Section S-V). We find that, for any choice of LV, replication efficiency reaches its maximum when the fractions of dimerization (1+1) reactions and erroneous V+V ligations are equal (Fig. 2C for LV = 6 nt). This criterion can be used to derive a scaling law for the optimal equilibrium concentration ratio as a function of the oligomer length LV, which is shown as a dashed line in Fig. 2F (the length-scale ΛF+F is defined in Supplementary Material Section S-V). The optimal equilibrium concentration ratio decreases exponentially with LV, while the hybridization energy γ/2 sets the inverse length-scale of the exponential decay. Analytical estimate and numerical solution agree well, as long as the hybridization is weak and oligomers are sufficiently short. For strong binding and long oligomers, complexes involving more than three strands play a non-negligible role, but such complexes are neglected in the analytical approximation.

Fig. 2G shows how the maximal replication efficiency depends on the VCG oligomer length. Consistent with the qualitative trend observed in Fig. 2D, longer oligomers enable higher maxima in replication efficiency. Regardless of the choice of γ, replication efficiency reaches 100% if LV is sufficiently high. Using Eq. (4), we find the following approximation for the maximal replication efficiency attainable at a given oligomer length (see dashed lines in Fig. 2G), where η° is a genome-dependent constant (Supplementary Material Section S-VI). This equation provides guidance for the construction of VCG pools with high replication efficiency: Given a target efficiency, the necessary oligomer length and hybridization energy, i.e., temperature, can be calculated. In Fig. 2G, we determine the oligomer length necessary to achieve ηmax = 95% for varying hybridization energies γ. At higher temperature (weaker binding), VCG pools require longer oligomers for replication with high efficiency.

Eq. (5) is not restricted to the example genome of length LG = 16 nt introduced previously, but may be applied to genomes of arbitrary length (provided the genomes follow the construction principles outlined in the method section). In such genomes, the minimal unique subsequence length scales logarithmically with the length of the genome, LS ∼ ln(LG). Based on LG and LS, we can compute the genome-dependent quantities ΛF+F and η° via the combinatorial multiplicities of productive complexes, which allow us to compute the characteristic VCG oligomer length at which the system reaches ηmax = 95%. Just like LS, this characteristic VCG oligomer length also increases logarithmically with the genome length (Fig. 2H), allowing VCG pools with moderate oligomer sizes to encode long circular genomes.

Pools containing VCG oligomers of different lengths are dominated by the long oligomers

In the previous section, we characterized the behavior of pools containing VCG oligomers of a single length. We observed that V+V ligations are error-prone due to insufficient overlap between the educt strands and the template, whereas F+V ligations extend the VCG oligomers with high efficiency. The F+V ligations will gradually broaden the length distribution of the VCG pool, raising the question of how this broadening affects the replication behavior. In principle, introducing multiple oligomer lengths into the VCG pool might even improve the fidelity of V+V ligations, since a long VCG oligomer could serve as a template for the correct ligation of two shorter VCG oligomers.

To analyze this question quantitatively, we first consider the simple case of a VCG pool that only contains monomers, tetramers, and octamers. The concentration of monomers is set to c(1) = 0.1 mM, while the concentrations of the VCG oligomers are varied independently (Fig. 3A). Replication efficiency reaches its maximum at c(8) ≈ 0.1 μM and very low tetramer concentration, c(4) ≈ 7.4 pM, effectively resembling a single-length VCG pool containing only octamers. As shown in Fig. 3B, the maximal efficiency is surrounded by a plateau of close-to-optimal efficiency. The octamer concentration can be varied by more than one order of magnitude without significant change in efficiency. Similarly, adding tetramers does not affect efficiency as long as the tetramer concentration does not exceed the octamer concentration. Fig. 3C illustrates that the plateau of close-to-optimal efficiency coincides with the concentration regime where the ligation of a monomer to an octamer with another octamer acting as template, , is the dominant ligation reaction. For high tetramer concentration and intermediate octamer concentration, templated ligation of tetramers on octamer templates, , surpasses the contribution of ligations (green shaded area in Fig. 3C). The reactions give rise to a ridge of increased efficiency, which represents a local maximum, but is small compared to the plateau of close-to-optimal efficiency (Fig. 3B). Even though reactions of the type produce correct products in most cases, they compete with error-prone ligations like or , which reduce fidelity. Increasing the binding affinity, i.e., lowering γ, enhances the contribution of ligations at the expense of ligations, resulting in an increased local maximum in replication efficiency (Supplementary Material Fig. S9). However, very strong hybridization energy, γ = −5.0 kBT, is necessary for ligations to give rise to similar efficiency as ligations. Thus, in this example, high efficiency replication facilitated by the ligation of short VCG oligomers on long VCG oligomers is in theory possible, but requires unrealistically strong binding affinity. Moreover, this mechanism is only effective if educts and template differ significantly with respect to their length. In case short and long oligomers are similar in size, templated ligation of two short oligomers tends to be error-prone, irrespective of the choice of binding affinity (Supplementary Material Fig. S10 and S11). In that case, F+V ligations remain the most reliable replication mechanism.

Replication performance of VCG pools containing VCG oligomers of multiple lengths (multi-length VCG pools). A The pool contains a fixed concentration of monomers, , as well as tetramers and octamers at variable concentration. The hybridization energy per nearest-neighbor block is γ = −2.5 kBT. B Replication efficiency reaches its maximum for c(8) ≈ 0.1 μM and significantly lower tetramer concentration, c(4) ≈ 7.4 pM. Efficiency remains close-to-maximal on a plateau around the maximum spanning almost two orders of magnitude in tetramer and octamer concentration. In addition, efficiency exhibits a ridge of increased efficiency for high tetramer concentration and intermediate octamer concentration. C The concentrations of complexes that facilitate templated ligation are grouped by the length of the template and the educts, . We distinguish complexes producing correct (labeled ”c”) and false products (labeled ”f”). For each relevant type of complex, we highlight the region of concentration where it contributes most significantly, i.e., at least 20% of the total ligation flux. The plateau of high efficiency is dominated by the ligation of monomers to octamers, whereas the ridge of increased efficiency is due to the correct ligation of two tetramers templated by an octamer.

We observe similar behavior in VCG pools containing a range of oligomer lengths from to (Fig. 4A), again using γ = −2.5 kBT (typical RNA hybridization energy). We maximize replication efficiency with respect to the concentration of VCG oligomers under the assumption that all VCG oligomers of any length have the same concentration (uniform profile). Efficiency again reaches its maximum when F+V ligations of long VCG oligomers are the predominant type of templated ligation (see Fig. 4D, which shows the relative contribution of different ligation types in VCG pools with fixed nt and varying nt). Adding short VCG oligomers to the pool leaves the relative contributions of the dominant ligation types almost unchanged and increases the fraction of unproductive F+F ligations only marginally. Replication efficiency is governed by the behavior of the long oligomers included in the pool. Decreasing for fixed reduces the achievable maximal replication efficiency only slightly. Conversely, lowering the maximal oligomer length, , while keeping fixed causes the efficiency to decrease significantly (Fig. 4B). This is plausible because, for any , replication efficiency is bounded by the replication efficiency of the longest oligomer in the pool (blue curve in Fig. 4B, which corresponds to the orange curve in Fig. 2G). In addition, for small , short and long VCG oligomers are increasingly similar in length, which gives rise to a broad spectrum of erroneous V+V ligations that compete with the error-free F+V ligations (Fig. 4E).

Replication performance of multi-length VCG pools. A The pool contains a fixed concentration of monomers, , as well as long oligomers in the range at variable concentration . The length dependence of the concentration profile is assumed to be uniform (for panels B, D, and E) or exponential (for panel C); its steepness is set by the parameter κV. B If the length distribution is uniform, reducing decreases the maximal efficiency, whereas increasing increases it. Pools containing a range of oligomer lengths are always outperformed by single-length VCGs (blue curve). C Assuming an exponential length distribution of VCG oligomers allows to tune from a poorly-performing regime (dominated by oligomers of length ) to a well-performing regime (dominated by oligomers of length ). In the limit κV → ∞, ηmax approaches the replication efficiency of single-length pools containing only oligomers of length (dashed lines). D Especially for high , replication is dominated by primer extension of the long oligomers in the VCG, here . In this limit, addition of shorter oligomers leaves the dominant F+V ligations almost unchanged. E Reducing for fixed nt gives rise to increased significance of erroneous ligation reactions.

In a realistic prebiotic setup, the concentration profile of the VCG pool would not be uniform. Depending on the mechanism that produces the pool and its coupling to the non-equilibrium environment, it might have a concentration profile that decreases or increases exponentially with length. We use the parameter κV to control this exponential length dependence (Fig. 4A): For negative κV, the concentration increases as a function of length, while exponentially decaying length distributions have positive κV. We find that replication efficiency is high if the concentration of long VCG oligomers exceeds or at least matches the concentration of short VCG oligomers (see κV ≤ 0 in Fig. 4C). In that case, replication efficiency is dominated by the long oligomers in the pool, since these form the most stable complexes. As the concentration of long oligomers is decreased further (κV > 0), the higher stability of complexes formed by longer oligomers is eventually insufficient to compensate for the reduced concentration of long oligomers. Replication efficiency is then governed by short VCG oligomers, which exhibit lower replication efficiency. In the limit κV → ∞, replication efficiency approaches the replication efficiency of a single-length VCG pool containing only oligomers of length (Fig. 4C and Fig. 2G).

Adding dinucleotides to the feedstock decreases the efficiency of replication

So far, we focused on ensembles that contain solely monomers as feedstock. However, examining the influence of dimers on replication in VCG pools is of interest, since dinucleotides have proven to be interesting candidates for enzyme-free RNA copying [9–11, 23]. For this reason, we study oligomer pools like those illustrated in Fig. 5A: The ensemble contains monomers, dimers and VCG oligomers of a single length, LV. As our default scenario, the dimer concentration is set to 10% of the monomer concentration, corresponding to κF = 2.3, but this ratio can be modified by changing κF.

Replication performance of single-length VCG pools containing monomers and dimers as feedstock. A The pool contains a fixed total concentration of feedstock oligomers, , partitioned into monomers and dimers, as well as VCG oligomers of a single length, LV. The proportion of monomers and dimers can be adjusted via κF, and the concentration of the VCG oligomers is a free parameter,. B Replication efficiency exhibits a maximum at intermediate VCG concentration in systems with (dashed blue curve) and without dimers (solid blue curve). The presence of dimers reduces replication efficiency significantly, as they enhance the ligation share of incorrect F+V ligations (dashed green curve). The panel depicts the behavior for LV = 7 nt and κF = 2.3. C Optimal replication efficiency increases as a function of oligomer length, LV, and asymptotically approaches a plateau (dashed lines, Eq. (6). The value of this plateau, , is determined by the competition between correct and false 2+V reactions, both of which grow exponentially with LV. Thus, depends on the relative concentration of the dimers in the pool: the more dimers are included, the lower is . D Erroneous 1+V ligations are possible if the educt oligomer has a short overlap region with the template. The hybridization energy for such configurations is small, and independent of the length of the VCG oligomers (left). 2+V ligations may produce incorrect products via the same mechanism (middle). In addition, erroneous 2+V ligations can be caused by complexes in which two VCG oligomers hybridize perfectly to each other, but the dimer has a dangling end. The stability of these complexes increases exponentially with oligomer length (right).

Fig. 5B compares the replication efficiency of a pool with and without dimers . In both cases, pools that are rich in VCG oligomers exhibit low efficiency of replication. As erroneous V+V ligations are the dominant type of reaction in this limit, the same efficiency is obtained regardless of the presence of dimers. In contrast, when the pool is rich in feedstock, pools with and without dimers behave differently: If only monomers are included, the efficiency approaches zero, as dimerizing monomers do not contribute to the yield, and thus not to the efficiency. However, the presence of dimers enables the ligation of monomers and dimers to form trimers and tetramers, which lead to a non-zero yield. Given the low VCG concentration, dimers are more abundant than VCG oligomers and thus more likely to act as templates. Therefore, educt oligomers can only hybridize to the template with a single nucleotide-long hybridization region. Consequently, many products are incompatible with the circular genome, causing low efficiency.

Replication efficiency reaches its maximum at intermediate VCG concentrations, where replication is dominated by F+V ligations. Notably, the maximal attainable efficiency is significantly lower for ensembles with dimers than without, as dimers increase the number and the stability of complex configurations that can form incorrect products (Fig. 5C). For , ligation products are only incorrect if the overlap between the VCG educt oligomer and template is shorter than the unique subsequence length, LS. For , however, dangling end dimers can cause incorrect products even in case of long overlap of educt oligomer and template (see right column in Fig. 5C). The stability of the latter complexes depends on the length of the VCG oligomers, LV, whereas the stability of complexes facilitating incorrect monomer addition is independent of oligomer length (Fig. 5C).

In the presence of dimers, the length-dependent stability of complexes facilitating correct and incorrect F+V ligations causes a competition, which sets an upper bound on the efficiency of replication (Supplementary Material Section S-VIII), Here, we introduced effective association constants 𝒦a, which depend differently on the VCG oligomer length, LV. While the effective association constant of complexes enabling incorrect 1+V ligations is length-independent, the effective association constant for incorrect 2+V ligations, , scales exponentially with LV, The effective association constants for correct 1+V and 2+V ligations also scale exponentially with the oligomer length (Supplementary Material Fig. S12).

In systems without dimers, i.e., κF → ∞, ηmax approaches 100%, which is consistent with the behavior observed in the previous sections. Conversely, in systems containing dimers, the maximal efficiency remains at a value below 100%, which depends on the concentration of dimers in the feedstock. Fig. 5C shows the dependence of maximal replication efficiency on the length of VCG oligomers in pools containing monomers, dimers and VCG oligomers. As LV increases, ηmax converges towards the upper bound defined in Eq. (6) (see dashed line in Fig. 5C).

Monomers preferentially ligate to short VCG oligomers in VCG-rich pools

Efficiency of replication is limited by the same mechanism regardless of the specifics of the underlying VCG pool: Oligomers that have an insufficiently long region of hybridization (shorter than LS) to their template are likely to produce an incorrect product upon ligation. In the previous sections, we minimized the contribution of such ligations by fine-tuning the concentration of VCG oligomers. Alternatively, incorrect templated ligations could be suppressed if only monomers are activated, while longer oligomers remain inactive. This scenario has already been investigated experimentally [18]. In a prebiotic setting, this scenario would emerge, for instance, if activated monomers are produced in the environment by an in situ activation chemistry, and subsequently transported into the compartment containing the VCG, while no activation takes place within the compartment [28, 29].

To explore the replication efficiency of this monomer-only activation scenario, we first consider a pool consisting only of activated monomers as well as non-activated dimers and VCG oligomers of a single length (Fig. 6A). We vary the concentration of VCG oligomers, but keep the feedstock concentrations constant. For small VCG concentrations, we observe low efficiencies (Fig. 6B), as the ligation of two monomers, or one monomer and one dimer, are most likely. Conversely, high facilitates the formation of complexes in which VCG oligomers are extended by monomers, which implies high efficiency (Fig. 6B). Note that, unlike in systems where all oligomers are activated, replication efficiency does not decrease for high VCG concentration, as erroneous V+V ligations are impossible. Instead, perfect replication efficiency (100%) is reached for sufficiently high LV.

Replication performance of single-length VCG pools in which only the monomers are activated. A The pool contains activated monomers alongside non-activated dimers and VCG oligomers of a single length. The concentrations of monomers and dimers are fixed, c(1) = 0.091 mM and c(2) = 9.1 μM, adding up to a total feedstock concentration of , while the concentration of VCG oligomers, , can be varied. B Unlike in pools in which all oligomers are activated, replication efficiency does not decrease at high VCG concentration if only monomers are activated, as erroneous V+V ligations are impossible. Instead, replication efficiency approaches an asymptotic value of 1. C The fraction of oligomers that are in a monomer extension-competent state depends on the total concentration of VCG oligomers. At low VCG concentration, most oligomers are single-stranded, and extension of oligomers by monomers is scarce. At high VCG concentration, r1+V approaches the asymptotic value (Eq. (7), grey dashed line). In this limit, almost all oligomers form duplexes, which facilitate monomer addition upon hybridization of a monomer. Thus, the asymptotic fraction of oligomers that gets extended by monomers is not determined by the oligomer length, but by the binding affinity of monomers to existing duplexes. Conversely, the threshold concentration at which depends on oligomer length (colored dashed lines): Longer oligomers reach higher r1+V at lower VCG concentration.

While replication efficiency characterizes the relative amount of nucleotides used for the correct elongation of VCG oligomers, we can also analyze which fraction of VCG oligomers is in a monomer extension-competent state, Here, denotes the equilibrium concentration of all complexes enabling the addition of a monomer to any VCG oligomer. We find that r1+V depends on the VCG concentration qualitatively in the same way as the efficiency: r1+V is small for small VCG concentration, butapproaches a value asymptotically for high VCGconcentration (Fig. 6C). In this limit, almost all VCG oligomers are part of a duplex. Monomers can bind to these duplexes to form complexes allowing for 1+V ligations. Since almost all VCG oligomers are already part of a duplex, increasing further does not increase the fraction of VCG oligomers that can be extended by monomers. Instead, the asymptotic value is determined by the concentration of monomers and their binding affinity Kd(1) to an existing duplex (Supplementary Material Section S-IX), In contrast to the more complex case considered further below, here, the asymptotic value does not depend on the length of the VCG oligomers. However, the threshold VCG concentration at which scales exponentially with LV (Supplementary Material Section S-IX), This scaling implies that longer oligomers require exponentially lower VCG concentration to achieve a given ratio r1+V (Fig. 6C), as their greater length allows them to form more stable complexes. This observation implies that pools with longer oligomers will always be more productive than pools with shorter oligomers (at equal VCG concentration).

The behavior becomes more complex in pools containing VCG oligomers of multiple lengths, due to the competitive binding within such heterogeneous pools. To illustrate this, we examine an ensemble containing VCG oligomers ranging from to along with the same feedstock as previously (activated monomers and non-activated dimers, see Fig. 7A). For simplicity, we assume that the length-distribution of VCG oligomers is uniform. We study the fraction of oligomers in a monomer extension-competent state as a function of oligomer length, where denotes the equilibrium concentration of all complexes that allow monomer-extension of VCG oligomers of length L. At low VCG concentration, longer oligomers are more likely to be extended by monomers than shorter ones (Fig. 7B). This behavior is intuitive, as longer oligomers tend to form more stable complexes, which lead to higher productivity. Surprisingly, increasing the VCG concentration reverses the length-dependence of the productivity, such that short oligomers are more likely to be extended by monomers than long ones (note the three crossings of the curves in Fig. 7B). For example, 8-mers are more likely to undergo primer-extension than 9-mers once the VCG concentration exceeds correct value ≈ 0.5 μM. We derived a semi-analytical expression for the threshold VCG concentrations at which oligomers of two different lengths have equal productivity (see dashed lines in Fig. 7B, Supplementary Material Sections S-X and S-XI).

Replication performance of multi-length VCG pools in which only the monomers are activated. A The pool contains activated monomers as well as non-activated dimers and VCG oligomers. The concentrations of monomers and dimers are fixed, c(1) = 0.091 mM and c(2) = 9.1 μM, adding up to a total feedstock concentration of , while the total concentration of VCG oligomers, , can be varied. All VCG oligomers are assumed to have the same concentration. B At low VCG concentration, the fraction of oligomers that are in a monomer extension-competent state is higher for long oligomers than for short oligomers, while, at high VCG concentration, monomers preferentially ligate to short oligomers (“inversion of productivity”). The threshold concentrations at which a short oligomer starts to outperform a longer oligomer depend on the lengths of the compared oligomers (dashed lines). C The mechanism underlying inversion of productivity can be understood based on the pair-wise competition of different VCG oligomers, e.g., 8-mers vs. 9-mers. Over the entire range of VCG concentrations, complexes with 8-mer templates have a lower relative equilibrium concentration than complexes with 9-mer templates (bottom two curves vs. top two curves). However, as the concentration of VCG oligomers is increased, the extension fraction of 8-mers that are extended by monomers using a 9-mer as a template exceeds the fraction of extended 9-mers. D The equilibrium concentration of free oligomer decreases with increasing . Long oligomers have a lower equilibrium concentration ratio of free oligomers, as they can form more stable complexes with longer hybridization sites. E Complexes in which 8-mers serve as template are less stable than complexes with 9-mer templates, explaining the higher relative equilibrium concentration of the latter complex type (see panel C). Complexes with 9-mer template have similar stability regardless of the length of the educt oligomer, i.e., and are similarly stable. This similar stability together with the higher concentration of free 8-mers compared to 9-mers (see panel D) is the reason why the fraction of monomer-extended 8-mers exceeds the one of 9-mers (see panel C).

To understand the mechanism underlying the inversion of productivity, we analyze how different complex types contribute to r1+V(L). We introduce which denotes the fraction of oligomers of length LE that are in a monomer extension-competent complex configuration that uses an template. The term oligomer of length LT as includes the sum over all possible configurations of complexes with the given lengths. Focusing on 8-mers and 9-mers as an example, we observe that complexes utilizing the 9-mer as template cause the inversion of productivity (top two curves in Fig. 7C): As the VCG concentration increases, the fraction of monomer-extendable 8-mers eventually surpasses the fraction of monomer-extendable 9-mers, i.e., (Fig. 7C). Two factors give rise to this feature: (i) Triplexes of type and have similar stability (Fig. 7E), and (ii) the equilibrium concentration of free 8-mers is higher than that of 9-mers (Fig. 7D). As a result, 8-mers are more likely than 9-mers to hybridize to an existing duplex , and, given the stability of the complex , 8-mers remain bound almost as stably as 9-mers. In summary, short oligomers sequester long oligomers as templates to enhance their monomer-extension rate, while long oligomers cannot make use of short oligomers as templates due to the relative instability of the corresponding complexes.

It is noteworthy that Ding et al. have already observed the inversion of productivity experimentally [18]. In their study, they included activated monomers, activated imidazolium-bridged dinucleotides and oligomers up to a length of 12 nt, and observed that the primer extension rate for short primers is higher than the extension rate of long primers. Even though our model differs from their setup in some aspects (e.g., different circular genome, no bridged dinucleotides) evaluating our model using parameters similar to those of the experimentally studied system predicts inversion of productivity that qualitatively agrees with the experimental findings (Supplementary Material Section S-XII). We therefore assume that the mechanism underlying inversion of productivity described here also applies to the experimental observations.

Discussion and conclusion

While significant progress has been made in understanding the prebiotic formation of (ribo)nucleotides [30–33] and characterizing ribozymes that might play a role in an RNA world [34–37], a convincing scenario bridging the gap between prebiotic chemistry and ribozyme-catalyzed replication is still missing. Here, we studied a scenario proposed by Zhou et al. [17] (the ‘Virtual Circular Genome’, VCG) using theoretical and computational approaches. We analyzed the process whereby template-directed ligation replicates the genomic information that is collectively stored in the VCG oligomers. Our analysis revealed a trade-off between the fidelity and the yield of this process: At low concentration of VCG oligomers, most of the ligations produce oligomers that are too short to specify a unique locus on the genome, resulting in a low yield of replication (Fig. 2B-C). At high VCG concentration, erroneous templated ligations limit the fidelity of replication (Fig. 2C-D). We considered two solutions to this trade-off: (i) a VCG pool composition that optimizes its replication behavior within the bounds of the fidelity-yield trade-off, and (ii) breaking the fidelity-yield trade-off by assuming that only monomers are activated.

The first solution maximizes the yield of replication for fixed fidelity. In pools containing only monomers and VCG oligomers of a single length, replication efficiency can be maximized by increasing the length of VCG oligomers and decreasing their concentration (Fig. 2F-G). This reduces the likelihood of error-prone templated ligation of long oligomers. When the pool contains VCG oligomers of multiple lengths, replication efficiency is typically governed by the longest oligomer in the pool (Fig. 4D). Including dimers as feedstock for the replication increases the error fraction (Fig. 5B-C), as dimers that bind to a template with a dangling end are prone to form an incorrect product (Fig. 5D).

The second solution eliminates the error-prone templated ligation of two VCG oligomers by assuming that only monomers are activated. This enables both fidelity and yield to remain high even at high VCG concentrations (Fig. 6B), effectively breaking the fidelity-yield trade-off. Unsurprisingly, longer VCG oligomers are then more likely to be extended than shorter oligomers at equal concentration, (Fig. 6C). However, this is only true for pools with VCG oligomers of a single length — once multiple VCG oligomer lengths compete with each other, shorter oligomers can be more productive than longer ones (Fig. 7B). This feature, which has also been observed experimentally [18], is caused by an asymmetry in the productive interaction between short and long oligomers (Fig. 7C): While short oligomers can sequester longer oligomers as templates for their extension by a monomer, short oligomers are unlikely to serve as templates for longer oligomers (Fig. 7D-E).

If the VCG scenario is to close the gap between prebiotic chemistry and ribozyme-catalyzed replication, VCG pools need to be capable of replicating (parts of) ribozymes that play a role in the emergence of life. While there are cases of small ribozymes [36] or ribozymes with small active sites, such as the Hammerhead ribozyme [38], ribozymes obtained experimentally via in vitro evolution are typically more than a hundred nucleotides long [4, 39–41]. In principle, long genomes (e.g., 1000 nt long) could be replicated in VCG pools containing relatively short oligomers (e.g., 10 nt long) with high efficiency (Fig. 2H). However, this requires a very special sequence composition of the genome. In our analysis, we assumed a clear separation between feedstock and VCG oligomers, defined by a single length scale LS, where all possible sequences of length L < LS are part of the feedstock and occur in the genome sequence, while any sequence of length L ≥ LS specifies a unique locus on the virtual genome. Such a sequence composition is particularly beneficial for the virtual circular genome scenario. However, the sequence composition of a generic genome sequence is characterized by two distinct length scales LS1 and LS2: (i) The former is defined as the largest length LS1 such that all possible subsequences of length L < LS1 are present in the genome, while (ii) LS2 is the smallest length such that all subsequences of the genome with length L ≥ LS2 are unique, i.e., they appear only once in the genome. These two length scales coincide for the specific sequence composition assumed by our analysis, but LS2 could be significantly longer than LS1 in prebiotically interesting genomes containing ribozyme sequences. Note that LS1 will generally scale logarithmically with the genome length, LS1 ∼ ln LG, while LS2 is constrained by the functional requirements of the biomolecules encoded by the genome.

For efficient replication of VCG pools, the second length scale, LS2, of the genome composition will determine the required length LV of the VCG oligomers. Our analysis showed that replication is efficient, if the VCG oligomers are significantly (more than two-fold) longer than LS, which is equal to LS2 for our idealized genome composition. We expect the interplay of the competing template-directed ligation processes in VCG pools in our idealized scenario to be transferable to the general case, such that genomes with arbitrary LS2 can also be efficiently replicated, if the oligomers in the pool are more than two-fold longer than LS2. These sufficiently long oligomers enable the formation of productive complexes, in which the hybridization regions are typically longer than LS2, leading to mostly correct products. On the contrary, if the characteristic length of oligomers in the VCG pool is too short, sequence scrambling leads to a loss of the information of the VCG [19].

Based on this reasoning, it is interesting to note that the virtual circular genome scenario creates a selection pressure for prebiotic genomes to reduce their LS2: A VCG pool encoding a genome containing multiple copies of certain long sequence motifs would replicate less efficiently than a pool for which a shorter motif length already specifies a unique locus on the genome. This comparison assumes all other properties of the two systems being the same, in particular the length distribution of the oligomers in the two pools. As our analysis shows, this length distribution is the key factor determining the replication efficiency for a given genome sequence. The length distribution, in turn, is determined by an interplay between the chemical kinetics of the system and the molecular transport controlled by its physical environment: For example, templated ligation in an open system with continuous in- and outflux of oligomers can lead to a non-monotoneous length distribution with a concentration peak at a characteristic oligomer length, Lc, which is set by the competition between dehybridization and outflux [20]. Via this emergent length-scale, the environment exerts a selection pressure on replication in the VCG scenario: While a VCG pool replicates its genome with high efficiency if the environment preferentially forms long oligomers, Lc > LS2, replication in environments with small characteristic oligomer length is expected to eliminate repeating sequence motifs that are longer than Lc. Mutational errors could replace long repeated sequence motifs by alternative sequences with shorter unique motifs, which may be functionally equivalent.

Given the broad range of prebiotically plausible non-equilibrium environments [42], it is reasonable to expect that some environments provide the required conditions for efficient replication. The constraints formulated in this work can help to guide the search for self-replicating VCG pools, in the vast space of possible concentration profiles and non-equilibrium environments.

eLife Assessment

This important theoretical study examines the possibility of encoding genomic information in a collective of short overlapping strands (e.g., the Virtual Circular Genome (VCG) model). The study presents solid theoretical arguments, simulations and comparisons to experimental data to point at potential features and limitations of such distributed collective encoding of information. The work should be of relevance to colleagues interested in molecular information processing and to those interested in pre-Central Dogma or prebiotic models of self-replication.

Reviewer #1 (Public review):

Summary:

This is an interesting theoretical study examining the viability of Virtual Circular Genome (VCG) model, a recently proposed scenario of prebiotic replication in which a relatively long sequence is stored as a collection of its shorter subsequences (and their compliments). It was previously pointed out that VCG model is prone to so-called sequence scrambling which limits the overall length of such a genome. In the present paper, additional limitations are identified. Specifically, it is shown that VCG is well replicated when the oligomers are elongated by sufficiently short chains from "feedstock" pool. However, ligation of oligomers from VCG itself results in a high error rate. I believe the research is of high quality and well written. However, the presentation could be improved and the key messages could be clarified.

(1) It is not clear from the paper whether the observed error has the same nature as sequence scrambling

(2) The authors introduce two important lengths LS1 and LS2 only in the conclusions and do not explain enough which each of them is important. It would make sense to discuss this early in the manuscript.

(3) It is not entirely clear why specific length distribution for VCG oligomers has to be assumed rather than emerged from simulations.

(4) Furthermore, the problem has another important length, L0 that is never introduced or discussed: a minimal hybridization length with a lifetime longer than the ligation time. From the parameters given, it appears that L0 is sufficiently long (~10 bases). In other words, it appears that the study is done is a somewhat suboptimal regime: most hybridization events do not lead to a ligation. Am I right in this assessment? If that is the case, the authors might want to explore another regime, L0

Strengths:

High-quality theoretical modeling of an important problem is implemented.

Weaknesses:

The conclusions are somewhat convoluted and could be presented better.

Reviewer #2 (Public review):

Summary:

This important theoretical and computational study by Burger and Gerland attempts to set environmental, compositional, kinetic, and thermodynamic constraints on the proposed virtual circular genome (VCG) model for the early non-enzymatic replication of RNA. The authors create a solid kinetic model using published kinetic and thermodynamic parameters for non-enzymatic RNA ligation and (de)hybridization, which allows them to test a variety of hypotheses about the VCG. Prominently, the authors find that the length (longer is better) and concentration (intermediate is better) of the VCG oligos have an outsized impact on the fidelity and yield of VCG production with important implications for future VCG design. They also identify that activation of only RNA monomers, which can be achieved using environmental separation of the activation and replication, can relax the constraints on the concentration of long VCG component oligos by avoiding the error-prone oligo-oligo ligation. Finally, in a complex scenario with multiple VCG oligo lengths, the authors demonstrate a clear bias for the extension of shorter oligos compared to the longer ones. This effect has been observed experimentally (Ding et al., JACS 2023) but was unexplained rigorously until now. Overall, this manuscript will be of interest to scientists studying the origin of life and the behavior of complex nucleic acid systems.

Strengths:

- The kinetic model is carefully and realistically created, enabling the authors to probe the VCG thoroughly.

- Fig. 6 outlines important constraints for scientists studying the origin of life. It supports the claim that the separation of activation and replication chemistry is required for efficient non-enzymatic replication. One could easily imagine a scenario where activation of molecules occurs, followed by their diffusion into another environment containing protocells that encapsulate a VCG. The selective diffusion of activated monomers across protocell membranes would then result in only activated monomers being available to the VCG, which is the constraint outlined in this work. The proposed exclusive replication by monomers also mirrors the modern biological systems, which nearly exclusively replicate by monomer extension.

- Another strength of the work is that it explains why shorter oligos extend better compared to the long ones in complex VCG mixtures. This point is independent of the activation chemistry used (it simply depends on the kinetics and thermodynamics of RNA base-pairing) so it should be very generalizable.

Weaknesses:

- Most of the experimental work on the VCG has been performed with the bridged 2-aminoimidazolium dinucleotides, which are not featured in the kinetic model of this work. Oher studies by Szostak and colleagues have demonstrated that non-enzymatic RNA extension with bridged dinucleotides have superior kinetics (Walton et al. JACS 2016, Li et al. JACS 2017), fidelity (Duzdevich et al. NAR 2021), and regioselectivity (Giurgiu et al. JACS 2017) compared to activated monomers, establishing the bridged dinucleotides as important for non-enzymatic RNA replication. Therefore, the omission of these species in the kinetic model presented here can be perceived as problematic. The major claim that avoidance of oligo ligations is beneficial for VCGs may be irrelevant if bridged dinucleotides are used as the extending species, because oligo ligations (V + V in this work) are kinetically orders of magnitude slower than monomer extensions (F + V in this work) (Ding et al. NAR 2022). Formally adding the bridged dinucleotides to the kinetic model is likely outside of the scope of this work, but perhaps the authors could test if this should be done in the future by simply increasing the rate of monomer extension (F + V) to match the bridged dinucleotide rate without changing rate of V + V ligation?

- The kinetic and thermodynamic parameters for oligo binding appear to be missing two potentially important components. First, base-paired RNA strands that contain gaps where an activated monomer or oligo can bind have been shown to display significantly different kinetics of ligation and binding/unbinding than complexes that do not contain such gaps (see Prywes et al. eLife 2016, Banerjee et al. Nature Nanotechnology 2023, and Todisco et al. JACS 2024). Would inclusion of such parameters alter the overall kinetic model? Second, it has been shown that long base-paired RNA can tolerate mismatches to an extent that can result in monomer ligation to such mismatched duplexes (see Todisco et al. NAR 2024). Would inclusion of the parameters published in Todisco et al. NAR 2024 alter the kinetic model significantly?

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

This is an interesting theoretical study examining the viability of Virtual Circular Genome (VCG) model, a recently proposed scenario of prebiotic replication in which a relatively long sequence is stored as a collection of its shorter subsequences (and their compliments). It was previously pointed out that VCG model is prone to socalled sequence scrambling which limits the overall length of such a genome. In the present paper, additional limitations are identified. Specifically, it is shown that VCG is well replicated when the oligomers are elongated by sufficiently short chains from ”feedstock” pool. However, ligation of oligomers from VCG itself results in a high error rate. I believe the research is of high quality and well written. However, the presentation could be improved and the key messages could be clarified.

Strengths:

High-quality theoretical modeling of an important problem is implemented.

Weaknesses:

The conclusions are somewhat convoluted and could be presented better.

(1) It is not clear from the paper whether the observed error has the same nature as sequence scrambling.

We thank the Reviewer for pointing out that this important point was not clearly explained. The sequence errors observed in our model are indeed of the same nature as sequence scrambling previously identified by Chamanian and Higgs (Chamanian and Higgs, PLoS Comp Biol 2022). The core issue is the ligation of two oligomers representing non-adjacent segments of the genome sequence, leading to the formation of ”chimeric” products that are not part of the desired genome.

Our analysis identifies the ligation of VCG oligomers (V+V reactions) as the primary mechanism driving sequence scrambling. This allowed us to propose two strategies to mitigate sequence scrambling: (i) tuning the length and concentration of the VCG oligomers, and (ii) considering scenarios where only feedstock monomers contribute to elongation (non-reactive VCG oligomers). We modified the Introduction and Results section of our manuscript to convey this connection more clearly.

(2) The authors introduce two important lengths LS1 and LS2 only in the conclusions and do not explain enough which each of them is important. It would make sense to discuss this early in the manuscript.

We agree with the Reviewer and have followed the suggestion to introduce the two important length scales earlier in the manuscript (in the Model section of the main text). In the updated version, we refer to these length scales as the exhaustive coverage length LE (formerly LS1) and the unique subsequence length LU (formerly LS2). The exhaustive coverage length LE is defined as the maximum motif length for which all possible sequences of that length appear somewhere in the genome. In contrast, the unique subsequence length LU is the minimum motif length such that each subsequence of that length occurs only once in the genome, thus giving each motif a unique ”address”.

Generally, a genome of length LG contains at most 2LG distinct subsequences, implying that LE can be at most , and LU must be at least , where ⌊...⌋ and ⌈...⌉ denote the next lower and higher integer, respectively. While the previous version of the manuscript focused exclusively on the limiting case LE = LmaxE and LU = LminU , we have extended our analysis to genomes with a broader range of LE and LU values the revised manuscript.

This extended analysis reveals that, for accurate and efficient replication, the VCG oligomer length must always exceed LU, regardless of the choice of LE. The required margin beyond LU depends on the distribution of intermediate-length motifs (i.e., with LE < L < LU), but is typically only a few nucleotides.

(3) It is not entirely clear why specific length distribution for VCG oligomers has to be assumed rather than emerged from simulations.

We have integrated these new findings into the Results section of the main text and expanded the discussion of their implications for the prebiotic relevance of the VCG scenario in the Discussion section. Full methodological details are provided in the Supplementary Material (Sections S1 and S8).

We thank the Reviewer for this insightful question. Our choice to assume specific length distributions for VCG oligomers is motivated by both conceptual and practical considerations. We explain our reasoning more clearly in the revised manuscript, in the beginning of the Model section of the main text.

Conceptually, our study focuses on the propagation of sequence information by an already-formed VCG, rather than its emergence from a random pool. As discussed by Chamanian and Higgs, the spontaneous formation of a VCG from randomly interacting oligomers is a rare event. Our aim is to understand whether, once formed, such a structure can robustly replicate under prebiotic conditions. This question is best addressed when the genome and the oligomer pool (including their lengths and concentrations) can be systematically controlled.

From a practical standpoint, working with a controllable pool of oligomers facilitates direct comparison to recent experimental studies that use predefined and well-characterized oligomer pools (Ding et al. JACS 2023). With our current methods and realistic rate constants, simulating the emergence of such pools from simple building blocks (e.g., monomers and dimers) would be computationally prohibitive, due to the low ligation rate. For example, in a system containing monomers (concentration 0.1mM) and octamers (concentration 1µM) in a volume of V = 3.3µm3, simulating the time between two ligation events takes over 300 hours of compute time (see SI Fig. S2). This renders dynamic pool generation unfeasible for the scope of our study.

(4) Furthermore, the problem has another important length, L0 that is never introduced or discussed: a minimal hybridization length with a lifetime longer than the ligation time. From the parameters given, it appears that L0 is sufficiently long (∼ 10 bases). In other words, it appears that the study is done is a somewhat suboptimal regime: most hybridization events do not lead to a ligation. Am I right in this assessment? If that is the case, the authors might want to explore another regime, _L_0 _< LS_1, by considering a higher ligation rate.

Indeed, we assume that the ligation rate is smaller than both the hybridization and dehybridization rates for any oligomer typically included in the pool (up to length 10). In terms of effective length scales, this corresponds to L0 ≈ 10nt, with L0 defined as stated by the Reviewer, i.e., the hybridization length corresponding to a lifetime comparable to the ligation time. Most of our analysis actually exploits the small ligation rate, by employing an adiabatic approximation in which ligation is assumed to be slower than any hybridization or dehybridization process in the pool irrespective of oligomer length. As the Reviewer states, in this regime most hybridization events are transient, and will not result in ligation, since the complexes typically dissociate before ligation can occur.

While we agree that this assumption limits the overall yield of replication, it has a beneficial effect on replication fidelity. Oligomers that hybridize with mismatches tend to unbind more quickly due to the destabilizing effect of mismatches. In the slow-ligation regime, such complexes are likely to dissociate before a ligation can occur, preventing the formation of incorrect products. In contrast, if the ligation rate was comparable to the unbinding rate of mismatched hybrids, these incorrect associations could undergo ligation, thereby lowering the fidelity of replication. We thus view the regime L0 > LV as more favorable for studying the error-suppressing potential of the VCG mechanism, though we acknowledge that exploring the effects of faster ligation rates is an interesting question for future work.

Reviewer #2 (Public review):

Summary:

This important theoretical and computational study by Burger and Gerland attempts to set environmental, compositional, kinetic, and thermodynamic constraints on the proposed virtual circular genome (VCG) model for the early non-enzymatic replication of RNA. The authors create a solid kinetic model using published kinetic and thermodynamic parameters for non-enzymatic RNA ligation and (de)hybridization, which allows them to test a variety of hypotheses about the VCG. Prominently, the authors find that the length (longer is better) and concentration (intermediate is better) of the VCG oligos have an outsized impact on the fidelity and yield of VCG production with important implications for future VCG design. They also identify that activation of only RNA monomers, which can be achieved using environmental separation of the activation and replication, can relax the constraints on the concentration of long VCG component oligos by avoiding the error-prone oligo-oligo ligation. Finally, in a complex scenario with multiple VCG oligo lengths, the authors demonstrate a clear bias for the extension of shorter oligos compared to the longer ones. This effect has been observed experimentally (Ding et al., JACS 2023) but was unexplained rigorously until now. Overall, this manuscript will be of interest to scientists studying the origin of life and the behavior of complex nucleic acid systems.

Strengths:

• The kinetic model is carefully and realistically created, enabling the authors to probe the VCG thoroughly.

• Fig. 6 outlines important constraints for scientists studying the origin of life. It supports the claim that the separation of activation and replication chemistry is required for efficient non-enzymatic replication. One could easily imagine a scenario where activation of molecules occurs, followed by their diffusion into another environment containing protocells that encapsulate a VCG. The selective diffusion of activated monomers across protocell membranes would then result in only activated monomers being available to the VCG, which is the constraint outlined in this work. The proposed exclusive replication by monomers also mirrors the modern biological systems, which nearly exclusively replicate by monomer extension.

• Another strength of the work is that it explains why shorter oligos extend better compared to the long ones in complex VCG mixtures. This point is independent of the activation chemistry used (it simply depends on the kinetics and thermodynamics of RNA base-pairing) so it should be very generalizable.

We thank the Reviewer for the careful assessment of our work and this concise summary of our main points.

Weaknesses:

• Most of the experimental work on the VCG has been performed with the bridged 2aminoimidazolium dinucleotides, which are not featured in the kinetic model of this work. Oher studies by Szostak and colleagues have demonstrated that non-enzymatic RNA extension with bridged dinucleotides have superior kinetics (Walton et al. JACS 2016, Li et al. JACS 2017), fidelity (Duzdevich et al. NAR 2021), and regioselectivity (Giurgiu et al. JACS 2017) compared to activated monomers, establishing the bridged dinucleotides as important for non-enzymatic RNA replication. Therefore, the omission of these species in the kinetic model presented here can be perceived as problematic. The major claim that avoidance of oligo ligations is beneficial for VCGs may be irrelevant if bridged dinucleotides are used as the extending species, because oligo ligations (V + V in this work) are kinetically orders of magnitude slower than monomer extensions (F + V in this work) (Ding et al. NAR 2022). Formally adding the bridged dinucleotides to the kinetic model is likely outside of the scope of this work, but perhaps the authors could test if this should be done in the future by simply increasing the rate of monomer extension (F + V) to match the bridged dinucleotide rate without changing rate of V + V ligation?

We thank the Reviewer for this insightful comment. Indeed, we did not design our model to specifically describe the use of bridged 2-aminoimidazolium dinucleotides as feedstock for the VCG scenario. Adding the bridged dinucleotides to our model would require allowing for feedstock that effectively changes its length during the ligation reaction. As anticipated already by the Reviewer, this is outside the scope of our current modeling framework, which was chosen to explore the generic issue of sequence scrambling in the VCG scenario without distinguishing between different types of activation chemistries.

Along the lines of the Reviewer’s suggestion, we clarified in the revised manuscript that we consider two limiting cases out of a family of models with two different ligation rate constants, klig,1 for ligations involving a monomer and klig,>1 for ligations involving no monomer, allowing for kinetic discrimination between these processes. We consider the two limiting cases where either klig,1 = klig,>1 or klig,1/klig,1 → 0. The latter case, captures the behavior expected from an activation chemistry that enables fast primer extension but slow ligation, thereby suppressing sequence scrambling via V+V ligation events. The corresponding results, presented in Figure 6 and 7, indeed show that the VCG replication efficiency approaches 100% for pools that are rich in VCG oligomers.

Our coarse-grained model, which does not explicitly describe the activation chemistry, was sufficient to capture important kinetic and thermodynamic constraints of the VCG scenario, and to qualitatively explain the experimental observation of a preferential extension of short over long VCG oligomers (Fig. 7B). For future work, we plan to extend our model to account for the activation chemistry in detail, to allow for a more quantitative comparison between theory and experiment.

• The kinetic and thermodynamic parameters for oligo binding appear to be missing two potentially important components. First, base-paired RNA strands that contain gaps where an activated monomer or oligo can bind have been shown to display significantly different kinetics of ligation and binding/unbinding than complexes that do not contain such gaps (see Prywes et al. eLife 2016, Banerjee et al. Nature Nanotechnology 2023, and Todisco et al. JACS 2024). Would inclusion of such parameters alter the overall kinetic model?

We thank the Reviewer for highlighting these recent studies. Todisco et al. (JACS 2024) report that complexes with gaps are well described by standard nearest-neighbor models, while stacking interactions at nick sites confer additional stability beyond these predictions. Our model is therefore expected to capture the thermodynamics of complexes with gaps accurately, but likely underestimates the stability of complexes containing nicks. In the VCG pool, all productive ligation complexes (F+F, F+V, V+V) inherently contain a nick and thus benefit from this stabilization, whereas unproductive complexes typically do not. The added stability is expected to increase the residence time of oligomers in productive complexes, thereby enhancing overall extension rates. However, since this stabilization applies uniformly across all productive complexes, it does not shift the relative contributions of different ligation pathways (in particular, correct vs. incorrect).

This reasoning assumes that hybridization and dehybridization occur on timescales faster than ligation or primer extension. It is conceivable that this separation of timescales does not hold, particularly for oligomers binding to templates with gaps, where association is slower due to steric hindrance, while dissociation is further slowed by stabilizing nicks. As a result, the residence time of such complexes can become comparable to (or longer than) the ligation timescale. We now discuss this aspect more thoroughly in the revised Results and Discussion sections. Capturing the resulting effects in our analytical framework would require relaxing the adiabatic assumption, which is beyond the scope of this work. We recognize the relevance of the non-adiabatic regime of the dynamics, and hope to explore this regime in follow-up work.

• Second, it has been shown that long base-paired RNA can tolerate mismatches to an extent that can result in monomer ligation to such mismatched duplexes (see Todisco et al. NAR 2024). Would inclusion of the parameters published in Todisco et al. NAR 2024 alter the kinetic model significantly?

In contrast to complexes with nicks and gaps, mismatched complexes (Todisco et al. NAR 2024) will decrease replication fidelity relative to the results presented in our manuscript. Our current model assumes perfect base pairing, such that replication errors arise only from binding events involving regions too short to reliably identify the correct genomic position (sequence scrambling). Allowing mismatches will indeed introduce an additional error mechanism via imperfect yet sufficiently stable duplexes, thereby increasing the rate of incorrect extensions. However, we expect this effect to be limited. Due to the thermodynamic cost of internal loops, mismatched duplexes most often have their mismatches near the ends of the hybridized region, where their destabilizing effect is weakest (Todisco et al. NAR 2024). Terminal mismatches at the 3’end of the primer have been shown to reduce the primer-extension rate significantly via a stalling effect (Rajamani et al. JACS 2010, Leu et al. JACS 2013). Hence, we would expect errors due to mismatched duplexes to primarily occur for mismatches at the 5′ end. Such errors could be mitigated by a VCG pool consisting only of oligomers that are sufficiently long relative to the unique motif length of the virtual genome.

We have extended the Discussion section to address this interesting issue.

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

• ’(apostrophes) should be prime symbols instead of apostrophes

We thank the Reviewer for spotting this mistake, which we have now corrected.

• In the Introduction, the section that discusses the fidelity of enzyme-free copying should include a reference to Duzdevich et al. NAR 2021, as that work measured the fidelity experimentally.

We have included this reference together with other references on the kinetics of hybridization/dehybridization to nicks and gaps in the main text.

• The term feedstock oligomers may be problematic, because these also include monomers. In the ”Templated ligation” section of the Model, the statement ”We consider pools in which all oligomers are activated, as well as pools in which only monomers are activated” is imprecise. ”All oligomers, including monomers,...” would be better so as to avoid confusion in readers accustomed to standard RNA language.

We thank the Reviewer for this helpful suggestion. In the revised manuscript, we now use the term feedstock (rather than feedstock oligomers) to avoid confusion. We have also revised the sentence in the ”Templated ligation” section to read ”all oligomers, including monomers, ...” as recommended.

• The ”Experimentally determined association rate constants” reference 24-26, which measured the rate constants for DNA. Considering that the authors are modeling RNA, I wonder if Ashwood et al. Biophysical Journal 2023 contains any relevant RNA data that could help refine the model?

We thank the Reviewer for pointing us to the study by Ashwood et al. We have added this reference to the corresponding paragraph in the revised manuscript. Their RNA association rate constant (∼ 5 × 107 M−1 s−1) is larger than the one we used (∼ 1×106 M−1 s−1), however a larger association rate is in fact beneficial for the validity of our adiabatic approximation, and thus would not affect our results, as long as the thermodynamic stability remains the same. This is because faster association then also implies faster dissociation, and the ratio of the ligation timescale to the timescales of (de)hybridization then becomes even smaller, which is the regime where the adiabatic approximation made in our analysis is justified.

• In ”Triplexe softype 1—8 and 1—9...”,the word triplexes will confuse readers with RNA expertise as triplexe simply a triple-strandedRNA.

We thank the Reviewer for pointing out the potentially ambiguous nomenclature. To avoid confusion with triplestranded RNA structures, we now refer to binary (ternary, ...) complexes instead of duplexes (triplexes, ...) throughout the revised manuscript.

Version 2

Abstract

Abstract

The transition from prebiotic chemistry to living systems requires the emergence of a scheme for enzyme-free genetic replication. Here, we analyze a recently proposed prebiotic replication scenario, the so-called Virtual Circular Genome (VCG) [Zhou et al., RNA 27, 1-11 (2021)]: Replication takes place in a pool of oligomers, where each oligomer contains a subsequence of a circular genome, such that the oligomers encode the full genome collectively. While the sequence of the circular genome may be reconstructed based on long oligomers, monomers and short oligomers merely act as replication feedstock. We observe a competition between the predominantly error-free ligation of a feedstock molecule to a long oligomer and the predominantly erroneous ligation of two long oligomers. Increasing the length of long oligomers and reducing their concentration decreases the fraction of erroneous ligations, enabling high-fidelity replication in the VCG. Alternatively, the formation of erroneous products can be suppressed if each ligation involves at least one monomer, while ligations between two long oligomers are effectively prevented. This kinetic discrimination (favoring monomer incorporation over oligomer–oligomer ligation) may be an intrinsic property of the activation chemistry, or can be externally imposed by selectively activating only monomers in the pool. Surprisingly, under these conditions, shorter oligomers are extended by monomers more quickly than long oligomers, a phenomenon which has already been observed experimentally [Ding et al., JACS 145, 7504-7515 (2023)]. Our work provides a theoretical explanation for this behavior, and predicts its dependence on system parameters such as the concentration of long oligomers. Taken together, the VCG constitutes a promising scenario of prebiotic information replication: It could mitigate challenges of in non-enzymatic copying via template-directed polymerization, such as short lengths of copied products and high error rates.

Introduction

In order to delineate possible pathways toward the emergence of life, it is necessary to understand how a chemical reaction network capable of storing and replicating genetic information might arise from prebiotic chemistry. RNA is commonly assumed to play a central role on this path, as it can store information in its sequence and catalyze its own replication (Joyce, 1989; Robertson and Joyce, 2012; Higgs and Lehman, 2015). While ribozymes capable of replicating strands of their own length have been demonstrated in the laboratory (Attwater et al., 2013), it remains elusive how enzyme-free self-replication might have worked before the emergence of such complex ribozymes.

One possible mechanism is template-directed primer extension (Kervio et al., 2016; Walton and Szostak, 2016; Ding et al., 2021; Leveau et al., 2022; Welsch et al., 2023): In this process, a primer hybridizes to a template and is extended by short oligonucleotides, thereby forming a (complementary) copy of the template strand. Considerable progress has been made in optimizing template-directed primer extension, but challenges remain: (i) The produced copies are likely to be incomplete. So far, at most 12 nt have been successfully added to an existing primer (Leveau et al., 2022). Moreover, as the pool of primer strands needs to emerge via random polymerization, the primer is likely to hybridize to the template at various positions, and not only to its 3′-end, leaving part of the 3′-end of the template uncopied (Szostak, 2011). (ii) Errors in enzyme-free copying are frequent due to the limited thermodynamic discrimination between correct Watson-Crick pairing and mismatches (Kervio et al., 2010; Leu et al., 2011, 2013). While some activation chemistries (relying on bridged dinucleotides) have been shown to exhibit improved fidelity (Duzdevich et al., 2021), the error probability still constrains the length of the genome that can be reliably replicated.

The issue of insufficient thermodynamic discrimination can, in principle, be mitigated by making use of kinetic stalling after the incorporation of a mismatch (Rajamani et al., 2010; Leu et al., 2013). By introducing a competition between the reduced polymerization rate and a characteristic timescale of the non-equilibrium environment, it is possible to filter correct sequences from incorrect ones (Göppel et al., 2021). To address the challenge of incomplete copies, Zhou et al. propose an new scenario of replication, the so-called Virtual Circular Genome (VCG) (Zhou et al., 2021). In this scenario, genetic information is stored in a pool of oligomers that are shorter than the circular genome they collectively encode: Each oligomer bears a subsequence of the circular genome, such that the collection of all oligomers encodes the full circular genome virtually. Within the pool, each oligomer can act as template or primer (Zhou et al., 2021). The oligomers hybridize to each other and form complexes that allow for templated ligation of two oligomers, or for the extension of an oligomer by a monomer. Because the sequences of the ligated strands and the template are part of the genome, most of the products should also retain the sequence of the genome. That way, long oligomers encoding the circular genome can be produced at the expense of short oligomers (Zhou et al., 2021). The long strands, in turn, can assemble into catalytically active ribozymes. With a continuous influx of short oligomers, the VCG might allow for continuous replication of the virtually encoded circular genome. Importantly, replication in the VCG is expected to avoid the issue of incomplete copies. Since the genome is circular, it does not matter which part of the genome an oligomer encodes, as long as the sequence is compatible with the genome sequence. An additional feature of the VCG scenario is that replication should be achievable without the need of adding many nucleotides to a primer: Provided the concentration of oligomers decreases exponentially with their length, the concentration of each oligomer in the pool can be doubled by extending each oligomer only by a few nucleotides (Zhou et al., 2021). The extension of an oligomer by a few nucleotides in a VCG pool has already been demonstrated experimentally (Ding et al., 2023).

A recent computational study points out that the VCG scenario is prone to loss of genetic information via “sequence scrambling” (Chamanian and Higgs, 2022): If the genome contains identical sequence motifs at multiple different loci, replication in the VCG will mix the sequences of these loci, thus destroying the initially defined genome. It is currently unclear, which conditions could prevent this genome instability of VCGs, such that their genetic information is retained. Length distribution, sequence composition, oligonucleotide concentration and environmental conditions, such as temperature, all affect the stability of complexes and thus the replication dynamics of the VCG pool. Here, we characterize the replication fidelity and yield of VCG pools using a kinetic model, which explicitly incorporates association and dissociation of RNA strands as well as templated ligation. We study a broad spectrum of prebiotically plausible and experimentally accessible oligomer pools, from pools containing only monomers and long oligomers of a single length to pools including a range of long oligomers with uniform or exponential concentration profile. The length of the included oligomers as well as their concentration are free parameters of our model.

We find that, regardless of the pool composition, three competing types of template-directed ligation reactions emerge: (i) ligations between two short oligomers (or monomers), producing products too short to specify a unique genomic locus, (ii) ligations between a short and a long oligomer, typically generating longer products compatible with the genome sequence, and (iii) ligations between two long oligomers, which often yield sequences incompatible with the genome. These erroneous ligation of type (iii) are a key driver of sequence scrambling, as they covalently link oligomers originating from non-adjacent genomic loci, effectively “mixing” distant regions of the genome. Fidelity is primarily determined by the competition between the correct extension of a long oligomer and the erroneous ligation of two long oligomers. The likelihood of the latter can be reduced by decreasing the relative abundance of long oligomers, even though this increases the frequency of unproductive ligations between short oligomers. As a result, fidelity can be improved at the cost of reduced yield. The efficiency, meaning the yield attainable at a fixed high fidelity, thus depends on the length distribution of the oligomers in the pool.

Alternatively, the issue of erroneous ligations is mitigated if ligations of long oligomers are kinetically suppressed, such that each ligation incorporates only one monomer at a time, as in the experimental study by Ding et al. (Ding et al., 2023). In this case, the VCG concentration can be chosen arbitrarily large without compromising fidelity. Interestingly, our model predicts an un-expected feature: In the limit of high VCG concentrations, short oligomers are more likely to be extended than long oligomers, even though, intuitively, complexes containing longer oligomers are expected to be more stable and thus more productive. The same behavior was indeed observed experimentally (Ding et al., 2023). We provide an explanation for this feature, and discuss its dependence on system parameters such as the length and the concentration of long oligomers in the pool.

Model

In the VCG scenario, a circular genome is stored in a pool of oligomers, with each oligomer shorter than the genome it helps encode. Each oligomer bears a subsequence of the circular genome, such that, collectively, the oligomers encode the full genome (Fig. 1A). As the spontaneous emergence of such VCG pools is expected to be rare (Chamanian and Higgs, 2022), our study focuses on the conditions under which an existing VCG pool can reliably replicate. We therefore begin with a known genome and an associated VCG pool, without addressing the question of origin. To set up our model of VCG dynamics, we specify (i) the circular genome used, (ii) the procedure by which the genome is mapped to a set of oligomers, and (iii) the chemical reactions governing the system’s evolution.

Model.

(A) In the Virtual Circular Genome (VCG) scenario, a circular genome (depicted in green) as well as its sequence complement are encoded in a pool of oligomers (depicted in blue and orange). Collectively, the pool of oligomers encodes the whole sequence of the circular genome. Depending on their length, two types of oligomers can be distinguished: Long VCG oligomers specify a unique locus on the genome, while feedstock molecules (monomers or short oligomers) are too short to do so. (B) The length-distribution of oligomers included in the VCG pool is assumed to be exponential. The concentration of feedstock and VCG oligomers as well as their respective length scales of exponential decay and can be varied independently. The set of included oligomer lengths can be restricted via and . (C) The hybridization energy of complexes is computed using a simplified nearest-neighbor model: Each full block comprised of two base pairs (depicted in pink) contributes γ, while dangling end blocks (depicted in blue) contribute γ/2. (D) Oligomers form complexes via hybridization reactions, or dehybridize from an existing complex. The ratio of hybridization and dehybridization rate is governed by the hybridization energy (Eq. 3). If two oligomers are adjacent to each other in a complex, they can undergo templated ligation. (E) Based on the length of the reacting oligomers, we distinguish three types of templated ligation: Ligation of two feedstock molecules (F+F), ligation of a feedstock molecule to a VCG oligomer (F+V) and ligation of two VCG oligomers (V+V).

Circular Genomes

For a given genome length, LG, there are distinct circular genomes (the factor 1/2LG accounts for the freedom to select the starting position and to choose the Watson or the Crick strand as reference sequence). A key property of the genome is its unique motif length, LU, defined as the shortest length such that all possible subsequences of length L ≥ LU appear at most once in the genome. In other words, all subsequences of length L ≥ LU specify a unique locus on the genome. In addition, each circular genome has another length scale, corresponding to the maximal motif length, up to which all possible motifs are contained in the genome. We refer to this length as the exhaustive coverage length, LE. We typically analyze unbiased genomes in which all possible subsequences of length L ≤ LE are contained at equal frequency.

Construction of VCG Pools

To specify a VCG pool that encodes a genomic sequence, one must select which subsequences are included in the pool at which concentrations. We consider unbiased pools, where the concentration of subsequences, c(L), depends only on their length, L, i.e., all subsequences of a given length are included at equal concentration. We refer to the length-dependent concentration profile as the length distribution of the pool. Depending on their length, oligomers fall into two categories (Fig. 1B): (i) short feedstock molecules (monomers and oligomers) and (ii) long VCG oligomers. Feedstock oligomers are oligomers that are shorter than the unique motif length LU. Since their sequence appears multiple times on the genome, they do not encode a specific position on the genome. Thus, they serve as feedstock for the elongation of VCG oligomers rather than as information storage. Conversely, VCG oligomers, which are longer than the unique motif length LU, have a unique locus on the circular genome. Collectively, the VCG oligomers enable the reconstruction of the full genome.

The full length distribution, c(L), can be decomposed into the contributions of feedstock and VCG oligomers, We assume that both cF and cV follow an exponential length distribution. In our model, the concentration of VCG oligomers can be varied independently of the concentration of feedstock, and the length scales for the exponential decay ( vs. ) may differ between feedstock and VCG oligomers. Additionally, we can restrict the set of oligomer lengths included in the pool by setting minimal and maximal lengths for feedstock and VCG oligomers individually, For any other oligomer length, the concentrations equal zero. This parametrization includes uniform length distributions as a special case (κF = 0 and κV = 0), and also allows for concentration profiles that are peaked. Peaked length distributions can emerge from the interplay of templated ligation, (de)hybridization and outflux in open systems (Rosenberger et al., 2021). We define the total concentration of feedstock, , as well as the total concentration of VCG oligomers, . Their ratio will turn out to be an important determinant of the VCG dynamics.

(De)hybridization Kinetics

Oligomers can hybridize to each other to form double-stranded complexes, or dehybridize from an existing complex. For simplicity, we do not include self-folding within a strand, which is a reasonable assumption for short oligomers. The stability of a complex is determined by its hybridization energy, with lower hybridization energy indicating greater stability. We use a simplified nearest-neighbor energy model to compute the hybridization energy (Rosenberger et al., 2021; Göppel et al., 2022; Laurent et al., 2024): The total energy equals the sum of the energy contributions of all nearest-neighbor blocks in a given complex (Fig. 1C). The energy contribution associated with a block of two Watson-Crick base pairs (matches) is denoted γ < 0, and dangling end blocks involving one Watson-Crick pair and one free base contribute γ/2. Nearest-neighbor blocks with mismatches increase the hybridization energy by γMM > 0 per block, thus reducing the stability of the complex. The rate constants of hybridization and dehybridization are related via where c° = 1 M is the standard concentration, and ΔG is the free energy of hybridization. The association rate constant kon is proportional to the encounter rate constant, kenc = 1/(c°t0). The encounter timescale t0 serves as the elementary time unit of the kinetic model, with all reaction timescales measured relative to it.

Templated Ligation

Two oligomers A and B that are hybridized adjacently to each other on a third oligomer C can produce a new oligomer A-B via templated ligation (Fig. 1D). Depending on the length of A and B, we distinguish three types of ligation reactions (Fig. 1E): (i) F+F ligations, in which two feedstock molecules ligate, (ii) F+V ligations, where a VCG oligomer is extended by a feedstock molecules, and (iii) V+V ligations involving two VCG oligomers. The formation of a covalent bond via templated ligation is not spontaneous, but requires the presence of an activation reaction. Usually, these reactions add a leaving group to the 5′-end of the nucleotide, which is cleaved during bond formation (Kervio et al., 2016; Walton and Szostak, 2016). We assume that the concentration of activating agent is sufficiently high for the activation to be far quicker than the formation of the covalent bond, such that activation and covalent bond formation can be treated as a single effective reaction. When not otherwise stated, we assume that all possible templated ligation reactions occur with the same rate constant klig.

Observables

Templated ligation in the pool forms longer oligomers at the expense of shorter oligomers and monomers. While the product of an F+V ligation (or V+V ligation) is always a VCG oligomer, F+F ligations can produce feedstock or VCG oligomers. In both cases, a produced VCG oligomer can be correct (compatible with the genome) or incorrect (incompatible). We quantify these processes by measuring extension fluxes in units of nucleotides ligated to an existing strand (counting the length of the shorter ligated strand as the number of incorporated bases). In particular, we define the fidelity f as the extension flux resulting in correct VCG oligomers relative to the flux resulting in any VCG oligomer, In addition, we introduce the yield y as the proportion of total extension flux that produces VCG oligomers, Efficient replication of the VCG requires both, high fidelity and high yield. Hence, we introduce the efficiency of replication η as the product of fidelity and yield, Moreover, we define the ligation share s of a ligation type, which allows us to discern the contributions of different types of templated ligations (F+F, F+V, V+V) to fidelity, yield, and efficiency,

Results

Replication efficiency reaches a maximum at intermediate concentrations of VCG oligomers

We begin our analysis of the dynamics of VCG pools with an exemplary genome of length LG = 16 nt, This genome contains all possible monomers and dimers with equal frequency, ensuring that all motifs up to LE = 2 nt are represented. Identifying a unique address on this genome requires at least three nucleotides. Therefore, the unique motif length is LU = 3 nt, and VCG oligomers need to be at least 3 nt long. Further below, we also explore genomes of different lengths LG, as well as varying characteristic sequence length scales LE and LU (method for genome construction detailed in Supplementary Section S1).

Based on the genome, we construct the initial oligomer pool. As a first step, we focus on a simple scenario in which the pool contains only monomers (serving as feedstock) and VCG oligomers of a single, defined length. The VCG pools are evolved in time using a stochastic simulation based on the Gillespie algorithm (Rosenberger et al., 2021; Göppel et al., 2022; Laurent et al., 2024). Since the Gillespie algorithm operates on the level of counts of molecules instead of concentrations, we must assign a volume to each system (in the range 1 µm3 to 10 000 µm3, see Supplementary Material Section S2). Besides the volume, we also need to choose the reaction rate constants appropriately: (i) The association time t0 is the fundamental time unit in our kinetic model, and all other times are expressed relative to t0. Experimentally determined association rate constants are typically around 106 −107 M−1 s−1 (Wetmur and Davidson, 1968; Braunlin and Bloomfield, 1991; Ashwood et al., 2023; Todisco et al., 2024a). For the purpose of estimating absolute timescales, we assume a constant association timescale of t0 ≈ 1 µs in the following. (ii) The timescale of dehybridization is computed via Eq. (3) using the energy contribution γ = −2.5 kBT for a matching nearest-neighbor block and γMM = 25.0 kBT in case of mismatches. The high energy penalty of nearest-neighbor blocks involving mismatches, γMM, is chosen to suppress the formation of mismatches, while the value of γ roughly matches the average energy of all matching nearest-neighbor blocks given by the Turner energy model of RNA hybridization (Mathews et al., 2004). (iii) For templated ligation, we select a reaction rate constant of . This choice of klig is consistent with ligation rates measured in enzyme-free template-directed primer extension experiments, which range from around (Sosson et al., 2019; Leveau et al., 2022) to roughly (Walton and Szostak, 2017), depending on the underlying activation chemistry. This indicates that, for sufficiently short oligomers (up to about 11 nt long), hybridization and dehybridization occur much faster than ligation, so binding equilibrates before ligation takes place.

Based on the ligation events observed in the simulation, we compute the observables introduced above (Model section). Due to the small ligation rate constant, it is computationally unfeasible to simulate the time evolution for more than a few ligation time units. Consequently, ligation events are scarce, which leads to high variances in the computed observables. We mitigate this issue by calculating the observables based on the concentration of complexes that are in a productive configuration, even if they do not undergo templated ligation within the time window of the simulation (Supplementary Material Section S2).

From the ligation events, we compute the total flux of oligomer formation. The observable y (yield) quantifies the fraction of this total flux directed to producing VCG oligomers. Fig. 2B shows how the yield depends on the concentration of VCG oligomers, , at a fixed total monomer concentration, . Here, the data points (with error bars) represent the simulation results with the different colors corresponding to different choices of initial VCG oligomer length, LV. We observe that the yield increases monotonically with the concentration of VCG oligomers, and approaches 100% for high . The concentration at which the pool reaches a yield of 50% depends on the oligomer length LV: Shorter VCG oligomers require higher concentrations for high yield. The concentration dependence of the yield can be rationalized by the types of templated ligations that are involved. For low VCG concentration, most templated ligation reactions are dimerizations (1+1) with the VCG oligomers merely acting as templates (Fig. 2C). As dimers are shorter than the unique motif length LU, their formation does not contribute to the yield, which explains the low yield in the limit of small . Conversely, for high VCG concentration, most of the templated ligations are F+V or V+V ligations, which produce oligomers of length L ≥ LU, implying high yield.

Replication performance of VCG pools containing VCG oligomers of a single length (single-length VCG pools).

(A) The pool contains a fixed concentration of monomers, , as well as VCG oligomers of a single length, L, at variable concentration (the VCG oligomers cover all possible subsequences of the genome and its complement at equal concentration). (B) The yield increases as a function of , because dimerizations become increasingly unlikely for high VCG concentrations. (C) The ligation share of different ligation types depends on the total VCG concentration: In the low concentration limit, dimerization (F+F) dominates; for intermediate concentrations, F+V ligations reach their maximum, while, for high concentrations, a substantial fraction of reactions are V+V ligations. The panel depicts the behavior for LV = 6 nt. (D) Replication efficiency is limited by the small yield for small . In the limit of high , replication efficiency decreases due to the growing number of error-prone V+V ligations. Maximal replication efficiency is reached at intermediate VCG concentration. (E) V+V ligations are prone to the formation of incorrect products due to the short overlap between educt strand and template. In general, the probability of correct product formation, pcorr, depends on the choice of circular genome and as well as its mapping to the VCG pool. The probabilities listed here refer to a VCG pool with LG = 16 nt, LE = 2 nt and LU = 3 nt. (F) The optimal equilibrium concentration ratio of free VCG strands to free feedstock strands, which maximizes replication efficiency, decays as a function of length (continuous line). The analytical scaling law (dashed line, Eq. (4)) captures this behavior. The window of close-to-optimal replication, within which efficiency deviates no more than 1% from its optimum (shaded areas), increases with LV, facilitating reliable replication without fine-tuning to match the optimal concentration ratio. (G) Maximal replication efficiency, which is attained at the optimal VCG concentration depicted in panel E, increases as a function of LV and approaches a plateau of 100%. For high efficiency, Eq. (5) provides a good approximation of the length-dependence of ηmax (dashed lines). The oligomer length at which replication efficiency equals 95% is determined using Eq. (5) (vertical dotted lines). (H) The unique motif length, , increases logarithmically with the length of the genome, LG. The length of VCG oligomers, LV, at which the optimal replication efficiency reaches 95% (computed using Eq. (5)) exhibits the same logarithmic dependence on LG.

Fig. 2C also shows that the relative contribution of V+V ligations increases with increasing , with a large fraction of them producing incorrect products (denoted V+V,f). This reduces the fidelity of replication, f, and causes a trade-off between fidelity and yield, which causes replication efficiency, η = f · y, to reach a maximum at intermediate VCG concentrations(Fig. 2D). Using hexamers as an example, Fig. 2E illustrates why V+V ligations are prone to forming incorrect oligomers: In order to ensure that two oligomers only ligate if their sequences are adjacent to each other in the true circular genome, both oligomers need to share an overlap of at least LU nucleotides with the template oligomer. Otherwise, the hybridization region is too short to identify the locus of the oligomer uniquely, and two oligomers from non-adjacent loci might ligate. The probability of forming incorrect products is a consequence of the combinatorics of possible subsequences in the VCG. Specifically, for a genome with LG = 16 nt, there are 32 different hexamers. For example, if the left educt hexamer only has one nucleotide of overlap with the template, there are 1/4 · 32 = 8 possible educt oligomers. However, only one out of those 8 hexamers is the correct partner for the right educt hexamer, implying an error probability of 1 − 1/8 = 7/8 (first example in Fig. 2E).

Characterizing replication efficiency via full simulations is computationally expensive: Depending on parameters, obtaining a single data point in Fig. 2B–D can require hundreds of simulations, each lasting several days. To explore a broad parameter space more easily, we introduce an approximate adiabatic method that (i) assumes ligation is much slower than any hybridization or dehybridization event, and (ii) relies on a coarse-grained sequence-independent representation of oligomers. Details are provided in Supplementary Material Sections S3 and S4. In brief, because ligation is rare, we first compute the equilibrium distribution of free and bound oligomers. Oligomers of the same length share a common concentration, and complex concentrations are determined via the mass action law using length-dependent dissociation constants. Combining the mass action law with a mass conservation constraint for each oligomer length allows us to compute the equilibrium concentrations of free VCG and feedstock strands, and , given total concentrations and . With these equilibrium values, we determine the concentrations of productive complexes and thus obtain the desired ligation-based observables without running full stochastic simulations.

The results of the adiabatic approach agree well with the simulation data (Fig. 2B-D), supporting that the replication efficiency depends non-monotonously on the concentration of VCG oligomers, with a maximum at intermediate concentration. While the available simulation data only allows for a qualitative characterization of this trend, the adiabatic approach enables a quantitative analysis. For instance, we use the adiabatic approach to determine the equilibrium concentration ratio at which replication efficiency is maximal as a function of the VCG oligomer length LV (solid line in Fig. 2F). As expected from the qualitative trend observed in the simulation, pools containing longer oligomers reach their maximum for lower concentration of VCG oligomers. The shaded area indicates the range of VCG concentrations within which the pool’s efficiency deviates by no more than one percent from its optimum. We observe that this range of close-to-optimal VCG concentrations increases with LV. Thus, pools containing longer oligomers require less fine-tuning of the VCG concentration for replication with high efficiency.

In addition to the numerical results, we utilize the adiabatic approach to study the optimal equilibrium VCG concentration analytically (Supplementary Material Section S6). We find that, for any choice of LV, replication efficiency reaches its maximum when the fractions of dimerization (1+1) reactions and erroneous V+V ligations are equal (Fig. 2C for LV = 6 nt). This criterion can be used to derive a scaling law for the optimal equilibrium concentration ratio as a function of the oligomer length LV, which is shown as a dashed line in Fig. 2F (the length-scale ΛF+F is defined in Supplementary Material Section S6). The optimal equilibrium concentration ratio decreases exponentially with LV, while the hybridization energy γ/2 sets the inverse length-scale of the exponential decay. Analytical estimate and numerical solution agree well, as long as the hybridization is weak and oligomers are sufficiently short. For strong binding and long oligomers, complexes involving more than three strands play a non-negligible role, but such complexes are neglected in the analytical approximation.

Fig. 2G shows how the maximal replication efficiency depends on the VCG oligomer length. Consistent with the qualitative trend observed in Fig. 2D, longer oligomers enable higher maxima in replication efficiency. Regardless of the choice of γ, replication efficiency reaches 100% if LV is sufficiently high. Starting from Eq. (4), we find the following approximation for the maximal replication efficiency attainable at a given oligomer length (dashed lines in Fig. 2G), where η° is a genome-dependent constant (Supplementary Material Section S7). This equation can provide guidance for the construction of VCG pools with high replication efficiency: Given a target efficiency, the necessary oligomer length and hybridization energy, i.e., temperature, can be calculated. In Fig. 2G, we determine the oligomer length necessary to achieve ηmax = 95% for varying hybridization energies γ. At higher temperature (weaker binding), VCG pools require longer oligomers to replicate with high efficiency.

Eq. (5) is not restricted to the specific example genome of length LG = 16 nt, but applies more generally to genomes of arbitrary length. Any genome of length LG can contain at most 2 LG distinct motifs. Consequently, the minimum length required to specify a unique address on the genome equals . By the same logic, the longest motif length for which all 4L possible sequences can be exhaustively represented within the genome is .Both of these characteristic lengths scale logarithmically with genome size. For genomes where LE is taken to be maximal and LU minimal, we find that the characteristic VCG oligomer length required for replication with high efficiency (ηmax = 95%) also scales logarithmically with genome length (Fig. 2H). Across genome lengths, the offset between LU and is roughly constant.

In genomes with different choices of LE and LU ( and ), the characteristic VCG oligomer length required for efficient replication, , is primarily determined by the unique motif length LU. Specifically, the length of VCG oligomers, LV, must exceed LU, regardless of the value of LE. This is shown for genomes of length LG = 64 nt in Supplementary Material Section S8, where we generate genomes with specified length scales LE and LU using a Metropolis–Hastings algorithm and analyze their replication efficiency. Intuitively, the required VCG oligomer length is set by LU, because LU defines the minimal hybridization region needed to ensure correct ligation via sequence-specific recognition. The precise offset between LU and depends on genome-specific features, such as the frequency distribution of motifs with lengths between LE and LU. If this distribution is nearly uniform (noting that at least one motif must repeat, or else it would constitute a unique address), then will be close to LU (Supplementary Material Fig. S9B). In contrast, strongly biased motif distributions require larger to achieve reliable replication (Supplementary Material Fig. S9A), though even in this case, typically exceeds LU by only a few nucleotides.

Multi-length VCG pools are dominated by longest oligomers

In the previous section, we characterized the behavior of pools containing VCG oligomers of a single length. We observed that V+V ligations are error-prone due to insufficient overlap between the educt strands and the template, whereas F+V ligations extend the VCG oligomers with high efficiency. The F+V ligations will gradually broaden the length distribution of the VCG pool, raising the question of how this broadening affects the replication behavior. In principle, introducing multiple oligomer lengths into the VCG pool might even improve the fidelity of V+V ligations, since a long VCG oligomer could serve as a template for the correct ligation of two shorter VCG oligomers.

To analyze this question quantitatively, we first consider the simple case of a VCG pool that only contains monomers, tetramers, and octamers. The concentration of monomers is set to c(1) = 0.1 mM, while the concentrations of the VCG oligomers are varied independently (Fig. 3A). Replication efficiency reaches its maximum at c(8) ≈ 0.1 µM and very low tetramer concentration, c(4) ≈ 7.4 pM, effectively resembling a single-length VCG pool containing only octamers. As shown in Fig. 3B, the maximal efficiency is surrounded by a plateau of close-to-optimal efficiency. The octamer concentration can be varied by more than one order of magnitude without significant change in efficiency. Similarly, adding tetramers does not affect efficiency as long as the tetramer concentration does not exceed the octamer concentration. Fig. 3C illustrates that the plateau of close-to-optimal efficiency coincides with the concentration regime where the ligation of a monomer to an octamer with another octamer acting as template, , is the dominant ligation reaction. For high tetramer concentration and intermediate octamer concentration, templated ligation of tetramers on octamer templates, , surpasses the contribution of ligations (green shaded area in Fig. 3C). The reactions give rise to a ridge of increased efficiency, which, however, is small compared to the plateau of close-to-optimal efficiency (Fig. 3B). Even though reactions of the type produce correct products in most cases, they compete with error-prone ligations like or , which reduce fidelity. Increasing the binding affinity, i.e., lowering γ, enhances the contribution of ligations at the expense of ligations, resulting in an increased local maximum in replication efficiency (Supplementary Material Fig. S11). However, very strong hybridization energy, γ = −5.0 kBT, is necessary for ligations to give rise to similar efficiency as ligations. Thus, in this example, high efficiency replication facilitated by the ligation of short VCG oligomers on long VCG oligomers is in theory possible, but requires unrealistically strong binding affinity. Moreover, this mechanism is only effective if educts and template differ significantly with respect to their length. If the involved VCG oligomers are too similar in size, templated ligation of two short oligomers tends to be error-prone, irrespective of the choice of binding affinity (Supplementary Material Fig. S12 and S13). In that case, F+V ligations remain the most reliable replication mechanism.

Replication performance of pools containing VCG oligomers of two different lengths.

(A) The pool contains a fixed concentration of monomers, , as well as tetramers and octamers at variable concentration. The hybridization energy per nearest-neighbor block is γ = −2.5 kBT. (B) Replication efficiency reaches its maximum for c(8) ≈ 0.1 µM and significantly lower tetramer concentration, c(4) ≈ 7.4 pM. Efficiency remains close-to-maximal on a plateau around the maximum spanning almost two orders of magnitude in tetramer and octamer concentration. In addition, efficiency exhibits a ridge of increased efficiency for high tetramer concentration and intermediate octamer concentration. (C) Complexes that facilitate templated ligation are grouped by the length of the template and the educts, . We distinguish complexes producing correct (labeled “c”) and false products (labeled “f”). For each relevant type of complex, we highlight the region in the concentration plane where it contributes most significantly, i.e., at least 20% of the total ligation flux. The plateau of high efficiency is dominated by the ligation of monomers to octamers, whereas the ridge of increased efficiency is due to the correct ligation of two tetramers templated by an octamer.

We observe similar behavior in VCG pools containing a range of oligomer lengths from to , rather than just tetramers and octamers. For a uniform VCG concentration profile (Fig. 4A, dark grey), the maximal replication efficiency (at the optimal VCG concentration) is attained when F+V ligations involving long VCG oligomers dominate the templated ligation (see Fig. 4D, showing the contribution of different ligation types in pools with varying at fixed , using γ = −2.5 kBT). Importantly, the maximal replication efficiency is always bounded by the efficiency of the longest VCG oligomer in the pool, independent of the presence or length of shorter oligomers (blue curve in Fig. 4B; identical to the orange curve in Fig. 2G). Including short VCG oligomers has minimal effect on the dominant ligation types and only slightly increases the proportion of unproductive F+F ligations. Consequently, reducing while keeping fixed leads to only a modest decline in maximal efficiency. In contrast, decreasing while holding constant causes a substantial reduction in efficiency (Fig. 4B), because the longest oligomer in the pool sets an upper bound on replication efficiency. Moreover, at low , short and long oligomers become more similar in length, giving rise to a spectrum of erroneous V+V ligations that compete with the productive F+V ligations (Fig. 4E).

Replication performance of multi-length VCG pools.

(A) The pool contains a fixed concentration of monomers, , as well as long oligomers in the range at variable concentration . The length dependence of the concentration profile is assumed to be uniform (for panels B, D, and E) or exponential (for panel C); its steepness is set by the parameter κV. (B) If the length distribution is uniform, reducing decreases the maximal efficiency, whereas increasing increases it. Pools containing a range of oligomer lengths are always outperformed by single-length VCGs (blue curve). (C) Assuming an exponential length distribution of VCG oligomers allows us to tune from a poorly-performing regime (dominated by oligomers of length ) to a well-performing regime (dominated by oligomers of length ). In the limit κV → ∞, ηmax approaches the replication efficiency of single-length pools containing only oligomers of length (dashed lines). (D) For high , replication is dominated by primer extension of the long oligomers in the VCG (here ). In this limit, addition of shorter oligomers leaves the dominant F+V ligations almost unchanged. (E) Reducing for fixed increases the fraction of unproductive (i.e., dimerization) or erroneous ligation reactions.

In a realistic prebiotic scenario, the concentration profile of the VCG pool would not be uniform. Depending on the mechanism producing the pool and its coupling to the non-equilibrium environment, it might have a concentration profile that decreases or increases exponentially with length. We use the parameter κV to control this exponential length dependence (Fig. 4A): For negative κV, the concentration increases as a function of length, while exponentially decaying length distributions have positive κV. We find that replication efficiency is high if the concentration of long VCG oligomers exceeds or at least matches the concentration of short VCG oligomers (κV ≤ 0 in Fig. 4C). In that case, replication efficiency is dominated by the long oligomers in the pool, since these form the most stable complexes. As the concentration of long oligomers is decreased further (κV > 0), the higher stability of complexes formed by longer oligomers is eventually insufficient to compensate for the reduced concentration of long oligomers. Replication efficiency is then governed by short VCG oligomers, which exhibit lower replication efficiency. In the limit κV → ∞, replication efficiency approaches the replication efficiency of a single-length VCG pool containing only oligomers of length (Fig. 4C and Fig. 2G).

Adding dinucleotides to the feedstock decreases replication efficiency

So far, we focused on ensembles that contain solely monomers as feedstock. However, examining the influence of dimers on replication in VCG pools is of interest, since dinucleotides have proven to be interesting candidates for enzyme-free RNA copying (Walton and Szostak, 2016; Sosson et al., 2019; Leveau et al., 2022). For this reason, we study oligomer pools like those illustrated in Fig. 5A: The ensemble contains monomers, dimers and VCG oligomers of a single length, LV. As our default scenario, the dimer concentration is set to 10% of the monomer concentration, corresponding to κF = 2.3, but this ratio can be modified by changing κF.

Replication performance of single-length VCG pools containing monomers and dimers as feedstock.

(A) The pool contains a fixed total concentration of feedstock, , partitioned into monomers and dimers, as well as VCG oligomers of a single length, LV. The proportion of monomers and dimers can be adjusted via κF, and the concentration of the VCG oligomers is a free parameter, . (B) Replication efficiency exhibits a maximum at intermediate VCG concentration in systems with (dashed blue curve) and without dimers (solid blue curve). The presence of dimers reduces replication efficiency significantly, as they enhance the ligation share of incorrect F+V ligations (dashed green curve). The panel depicts the behavior for LV = 7 nt and κF = 2.3. (C) Optimal replication efficiency increases as a function of oligomer length, LV, and asymptotically approaches a plateau (dashed lines, Eq. (6)). The value of this plateau, , is determined by the competition between correct and false 2+V reactions, both of which grow exponentially with LV. Thus, depends on the relative concentration of the dimers in the pool: the more dimers are included, the lower is . (D) Erroneous 1+V ligations are possible if the educt oligomer has a short overlap region with the template. The hybridization energy for such configurations is small, and independent of the length of the VCG oligomers (left). While 2+V ligations may produce incorrect products via the same mechanism (middle), they can also be caused by complexes in which two VCG oligomers hybridize perfectly to each other, but the dimer has a dangling end. The stability of these complexes increases exponentially with oligomer length (right).

Fig. 5B compares the replication efficiency of a pool with and without dimers . In both cases, pools that are rich in VCG oligomers exhibit low efficiency of replication. As erroneous V+V ligations are the dominant type of reaction in this limit, all pools achieve the same efficiency regardless of the presence of dimers. In contrast, when the pool is rich in feedstock (small ), pools with and without dimers behave differently: If only monomers are included, the efficiency approaches zero, as dimerizing monomers do not contribute to the yield, and thus not to the efficiency. However, the presence of dimers enables the ligation of monomers and dimers to form trimers and tetramers, which lead to a non-zero yield. Given the low VCG concentration, the ligations are likely to proceed using dimers as template. As a consequence, educt oligomers can only hybridize to the template with a single-nucleotide-long hybridization region, leading to frequent formation of incorrect products and, consequently, low efficiency.

Replication efficiency reaches its maximum at intermediate VCG concentrations, where replication is dominated by F+V ligations. Notably, the maximal attainable efficiency is significantly lower for pools with dimers than without, as dimers increase the number and the stability of complex configurations that can form incorrect products (Fig. 5C). Without dimers, ligation products are only incorrect if the overlap between the VCG educt oligomer and template is shorter than the unique motif length, LU. With dimers, however, dangling end dimers can cause incorrect products even in case of long overlap of educt oligomer and template (right column in Fig. 5C). The stability of the latter complexes depends on the length of the VCG oligomers, LV, whereas the stability of complexes facilitating incorrect monomer addition is independent of oligomer length (Fig. 5C).

In the presence of dimers, the length-dependent stability of complexes allowing for correct and incorrect F+V ligations causes a competition, which sets an upper bound on the efficiency of replication (Supplementary Material Section S10), Here, we introduced effective association constants 𝒦a, which depend differently on the VCG oligomer length, LV. While the effective association constant of complexes enabling incorrect 1+V ligations is length-independent, the effective association constant for incorrect 2+V ligations, , scales exponentially with LV, The effective association constants for correct 1+V and 2+V ligations also scale exponentially with the oligomer length (Supplementary Material Fig. S14).

In systems without dimers, i.e., 𝒦F → ∞, ηmax approaches 100%, which is consistent with the behavior observed in the previous sections. Conversely, in systems containing dimers, the maximal efficiency remains at a value below 100%, which depends on the concentration of dimers in the feedstock. Fig. 5C shows the dependence of maximal replication efficiency on the length of VCG oligomers in pools containing monomers, dimers and VCG oligomers. As LV increases, ηmax converges towards the upper bound defined in Eq. (6) (dashed line in Fig. 5C).

Kinetic suppression of error-prone VCG oligomer ligation

In all scenarios considered so far, the efficiency of replication is limited by a common mechanism, regardless of the specifics of the VCG pool: Ligations involving two oligomers that hybridize to the template over a region shorter than LU are prone to generate incorrect products. In previous sections, we minimized these erroneous ligations by fine-tuning the concentration and length of VCG oligomers. However, such control may become unnecessary if the typically error-prone ligation of two oligomers (V+V) is kinetically suppressed. Kinetic suppression can be an intrinsic property of the activation chemistry: Templated ligation of two oligomers can be several orders of magnitude slower than the extension of an oligomer by a single monomer (Prywes et al., 2016; Ding et al., 2021). As a result, V+V ligations are disfavored purely by their slower kinetics. In addition, it is conceivable that only monomers are chemically activated while longer oligomers remain inactive, which would further reduce the likelihood of erroneous ligations. This scenario has already been explored experimentally (Ding et al., 2023). In natural environments, it could occur, for instance, when activated monomers are produced externally and then diffuse into compartments containing the VCG but lacking internal activation pathways (Toparlak et al., 2023; Kriebisch et al., 2024).

Within our model, we capture the kinetic suppression by introducing two different rates of ligation, klig,1 for ligations involving a monomer and klig,>1 for ligations involving no monomer, allowing for kinetic discrimination between these processes. We explore the resulting replication efficiency in the limit of perfect kinetic discrimination (klig,>1/klig,1 → 0) where only monomers are reactive for ligation. We first consider a pool where the reactive monomers are mixed with VCG oligomers of a single length as well as non-reactive dimers (Fig. 6A). We vary the concentration of VCG oligomers, but keep the feedstock concentrations constant. For small VCG concentrations, we observe low efficiencies (Fig. 6B), as the ligation of two monomers, or one monomer and one dimer, are most likely. Conversely, high facilitates the formation of complexes in which VCG oligomers are extended by monomers, which implies high efficiency (Fig. 6B). Note that, unlike in systems where all oligomers are reactive, replication efficiency does not decrease for high VCG concentration, as erroneous V+V ligations are impossible. Instead, perfect replication efficiency (100%) is reached for sufficiently high LV.

Replication performance of single-length VCG pools with kinetic suppression of ligation between oligomers.

(A) The pool contains reactive monomers alongside non-reactive dimers and VCG oligomers of a single length. The concentrations of monomers and dimers are fixed, c(1) = 0.091 mM and c(2) = 9.1 µM, adding up to a total feedstock concentration of , while the concentration of VCG oligomers, , is varied. (B) Unlike in pools in which all ligation processes occur, replication efficiency does not decrease at high VCG concentration if ligations not involving monomers are kinetically suppressed. Instead, replication efficiency approaches an asymptotic value of 100%, as erroneous V+V ligations are impossible. (C) The fraction of oligomers that are in a monomer-extension-competent state depends on the total concentration of VCG oligomers. At low VCG concentration, most oligomers are single-stranded, and extension of oligomers by monomers is scarce. At high VCG concentration, r1+V approaches the asymptotic value (Eq. (7), grey dashed line). In this limit, almost all oligomers form duplexes, which facilitate monomer addition upon hybridization of a monomer. Thus, the asymptotic fraction of oligomers that gets extended by monomers is not determined by the oligomer length, but by the binding affinity of monomers to existing duplexes. Conversely, the threshold concentration at which depends on oligomer length (colored dashed lines): Longer oligomers reach higher r1+V at lower VCG concentration.

While replication efficiency characterizes the relative amount of nucleotides used for the correct elongation of VCG oligomers, it is also interesting to analyze which fraction of VCG oligomers is in a monomer-extension-competent state, Here, denotes the equilibrium concentration of all complexes enabling the addition of a monomer to any VCG oligomer. We find that r1+V depends on the VCG concentration qualitatively in the same way as the efficiency: r1+V is small for small VCG concentration, but approaches a value asymptotically for high VCG concentration (Fig. 6C). In this limit, almost all VCG oligomers are part of a duplex. Monomers can bind to these duplexes to form complexes allowing for 1+V ligations. Since almost all VCG oligomers are already part of a duplex, increasing further does not increase the fraction of VCG oligomers that can be extended by monomers. Instead, the asymptotic value is determined by the concentration of monomers and their binding affinity Kd (1) to an existing duplex (Supplementary Material Section S11), While the asymptotic value does not depend on the length of the VCG oligomers, the threshold VCG concentration at which scales exponentially with LV (Supplementary Material Section S11), This scaling implies that longer oligomers require exponentially lower VCG concentration to achieve a given ratio r1+V (Fig. 6C), as their greater length allows them to form more stable complexes. This observation implies that pools with longer oligomers will always be more productive than pools with shorter oligomers (at equal VCG concentration).

The behavior becomes more complex in pools containing VCG oligomers of multiple lengths, due to the competitive binding within such heterogeneous pools. To illustrate this, we examine an ensemble containing VCG oligomers ranging from to , along with the same feedstock as previously (reactive monomers and non-reactive dimers, see Fig. 7A). For simplicity, we assume that the length-distribution of VCG oligomers is uniform. We study the fraction of oligomers in a monomer-extension-competent state as a function of oligomer length, where denotes the equilibrium concentration of all complexes that allow monomer-extension of VCG oligomers of length L. At low VCG concentration, longer oligomers are more likely to be extended by monomers than shorter ones (Fig. 7B). This behavior is intuitive, as longer oligomers tend to form more stable complexes, which lead to higher productivity. Surprisingly, increasing the VCG concentration reverses the length-dependence of the productivity, such that short oligomers are more likely to be extended by monomers than long ones (note the three crossings of the curves in Fig. 7B). For example, 8-mers are more likely to undergo primer-extension than 9-mers once the VCG concentration exceeds ≈ 0.5 µM. We derived a semi-analytical expression for the threshold VCG concentrations at which oligomers of two different lengths have equal productivity (dashed lines in Fig. 7B, Supplementary Material Sections S12 and S13).

Replication performance of multi-length VCG pools with kinetic suppression of ligation between oligomers.

(A) The pool contains reactive monomers as well as non-reactive dimers and VCG oligomers. The concentrations of monomers and dimers are fixed, c(1) = 0.091 mM and c(2) = 9.1 µM, adding up to a total feedstock concentration of , while the total concentration of VCG oligomers, , is varied. All VCG oligomers are assumed to have the same concentration. (B) At low VCG concentration, long oligomers are more likely in a monomer-extension-competent state than short oligomers, whereas at high VCG concentration, the trend reverses and short oligomers are more likely to be extended by monomers (“inversion of productivity”). The threshold concentration at which a short oligomer starts to outperform a longer oligomer depends on the lengths of the compared oligomers (dashed lines). (C) The mechanism underlying inversion of productivity can be understood based on the pair-wise competition of different VCG oligomers, e.g., 8-mers vs. 9-mers. Over the entire range of VCG concentrations, complexes with 8-mer templates have a lower relative equilibrium concentration than complexes with 9-mer templates (bottom two curves vs. top two curves). As the concentration of VCG oligomers is increased, ligations of type exceed ligations of type , i.e., the extension fraction of 8-mers that are extended by monomers using a 9-mer as a template exceeds the fraction of extended 9-mers. (D) The equilibrium concentration of free oligomer decreases with increasing . For longer oligomers, the equilibrium fraction of free oligomers is lower, as they can form more stable complexes with longer hybridization sites. (E) Complexes in which 8-mers serve as template are less stable than complexes with 9-mer templates, explaining why complexes with 8-mer template are more abundant than complexes with 9-mer template (see panel C). Complexes with 9-mer template have similar stability regardless of the length of the educt oligomer, i.e., and are similarly stable. This similar stability together with the higher concentration of free 8-mers compared to 9-mers (see panel D) is the reason why the fraction of monomer-extended 8-mers exceeds the one of 9-mers (see panel C).

To understand the mechanism underlying the inversion of productivity, we analyze how different complex types contribute to r1+V(L). We introduce which denotes the fraction of oligomers of length LE that are in a monomer-extension-competent complex configuration that uses an oligomer of length LT as template. The term includes the sum over all possible configurations of complexes with the given lengths. Focusing on 8-mers and 9-mers as an example, we observe that complexes utilizing the 9-mer as template are responsible for the inversion of productivity (top two curves in Fig. 7C): As the VCG concentration increases, the fraction of monomer-extendable 8-mers eventually surpasses the fraction of monomer-extendable 9-mers, i.e., (Fig. 7C). Two factors give rise to this feature: (i) Ternary complexes of type and have similar stability (Fig. 7E), and (ii) the equilibrium concentration of free 8-mers is higher than that of 9-mers (Fig. 7D). As a result, 8-mers are more likely than 9-mers to hybridize to an existing duplex , and, given the stability of the complex , 8-mers remain bound almost as stably as 9-mers. In summary, short oligomers sequester long oligomers as templates to enhance their monomer-extension rate, while long oligomers cannot make use of short oligomers as templates due to the relative instability of the corresponding complexes.

It is noteworthy that Ding et al. have already observed the inversion of productivity experimentally (Ding et al., 2023). In their study, they included activated monomers, activated imidazolium-bridged dinucleotides and oligomers up to a length of 12 nt, and observed that the primer extension rate for short primers is higher than the extension rate of long primers. Even though our model differs from their setup in some aspects (e.g., different circular genome, no bridged dinucleotides) evaluating our model using parameters similar to those of the experimentally studied system predicts inversion of productivity that qualitatively agrees with the experimental findings (Supplementary Material Section S14). We therefore assume that the mechanism underlying inversion of productivity described here also applies to the experimental observations.

Discussion and Conclusion

While significant progress has been made in understanding the prebiotic formation of ribonucleotides (Powner et al., 2009; Kim et al., 2011; Benner et al., 2012; Becker et al., 2016) and characterizing ribozymes that might play a role in an RNA world (Mutschler et al., 2015; Attwater et al., 2018; Pressman et al., 2019; Tjhung et al., 2020), a convincing scenario bridging the gap between prebiotic chemistry and ribozyme-catalyzed replication is still missing. Here, we studied a scenario proposed by Zhou et al. (Zhou et al., 2021) (the ‘Virtual Circular Genome’, VCG) using theoretical and computational approaches. We analyzed the process whereby template-directed ligation replicates the genomic information that is collectively stored in the VCG oligomers. Our analysis revealed a trade-off between the fidelity and the yield of this process: At low concentration of VCG oligomers, most of the ligations produce oligomers that are too short to specify a unique locus on the genome, resulting in a low yield of replication (Fig. 2B-C). At high VCG concentration, erroneous templated ligations cause sequence scrambling and consequently limit the fidelity of replication (Fig. 2C-D). We considered two solutions to these issues: (i) a VCG pool composition that optimizes its replication behavior within the bounds of the fidelity-yield trade-off, and (ii) breaking the fidelity-yield trade-off given that error-prone ligations can be kinetically suppressed.

The first solution maximizes the yield of replication for fixed fidelity. In pools containing only monomers and VCG oligomers of a single length, replication efficiency can be maximized by increasing the length of VCG oligomers and decreasing their concentration (Fig. 2F-G). This reduces the likelihood of error-prone templated ligation of long oligomers. When the pool contains VCG oligomers of multiple lengths, replication efficiency is typically governed by the longest oligomer in the pool (Fig. 4D). Including dimers as feedstock for the replication increases the error fraction (Fig. 5B-C), as dimers that bind to a template with a dangling end are prone to form an incorrect product (Fig. 5D).

The second solution eliminates the error-prone templated ligation of two VCG oligomers by suppressing them kinetically, e.g., by assuring that only monomers are chemically activated. This enables both fidelity and yield to remain high at high VCG concentrations (Fig. 6B), effectively breaking the fidelity-yield trade-off. Longer VCG oligomers are then more likely to be extended than shorter oligomers at equal concentration, (Fig. 6C). However, this is only true for pools with VCG oligomers of a single length — once multiple VCG oligomer lengths compete with each other, shorter oligomers can be more productive than longer ones (Fig. 7B). This feature, which has also been observed experimentally (Ding et al., 2023), is caused by an asymmetry in the productive interaction between short and long oligomers (Fig. 7C): While short oligomers can sequester longer oligomers as templates for their extension by a monomer, short oligomers are unlikely to serve as templates for longer oligomers (Fig. 7D-E).

As we intended to study the pathways responsible for sequence scrambling and to explore possible mitigation strategies, we based our analysis on a coarse-grained model which neglects some experimental details. First, we assumed that a complex instantaneously dehybridizes if it contains a non-complementary base pair, whereas in reality, short duplexes can tolerate a limited number of mismatches (Todisco et al., 2024a). While such mismatches can facilitate incorrect hybridization and introduce additional replication errors, we expect this effect to be moderate: Mismatches preferentially occur near the ends of the hybridized region, where their destabilizing effect on binding is weakest (Todisco et al., 2024a). However, such terminal mismatches have also been shown to significantly reduce ligation rates (Rajamani et al., 2010; Leu et al., 2013), which in turn limits the likelihood of forming incorrect products.

Second, we simplified the hybridization dynamics by assuming that all oligomers bind to each other at equal rates, and that dehybridization rates are determined by the hybridization energy computed via a nearest-neighbor model. However, recent work has shown that hybridization to a gap flanked by two oligomers proceeds more slowly than binding to an unoccupied template. Moreover, the resulting nicked complexes (two oligomers hybridized adjacently on a template) are more stable than predicted by standard nearest-neighbor models due to enhanced stacking interactions at the nick site (Todisco et al., 2024b). While this added stability is not expected to affect overall replication efficiency of the VCG (since all productive complexes, correct or incorrect, contain a nick), it can impact the kinetics of the system. In particular, the extended lifetime of such complexes may challenge the adiabatic approximation used in much of our analysis, which assumes ligation is always slower than hybridization and dehybridization.

Third, we do not model the activation chemistry explicitly, but instead assume that all monomers (and, depending on the scenario, also oligomers) are always reactive. As a result, some activated intermediates that are known to form in experiments, such as imidazolium-bridged dinucleotides (Walton and Szostak, 2016), are not modeled. Nonetheless, we include aspects of activation chemistry in a coarse-grained manner. Specifically, to capture the experimentally observed difference in reactivity between monomer incorporation and templated ligation of oligomers under aminoimidazolium activation, we introduce two distinct ligation rate constants. With this approach, we describe the experimental setup well enough to qualitatively reproduce features observed in experiments, for example, the preferential extension of shorter oligomers by monomers in pools containing VCG oligomers of varying lengths (Ding et al., 2023).

The VCG scenario was proposed to close the gap between prebiotic chemistry and ribozyme-catalyzed replication. To this end, VCG pools need to be capable of replicating (parts of) ribozymes that play a role in the emergence of life. While there are cases of small ribozymes (Pressman et al., 2019) or ribozymes with small active sites (e.g., the Hammerhead ribozyme (Scott et al., 2013)), ribozymes obtained experimentally via in vitro evolution are often more than a hundred nucleotides long (Johnston et al., 2001; Müller and Bartel, 2008; Wochner et al., 2011; Attwater et al., 2013). Remarkably, our model suggests that the VCG scenario enables high-fidelity replication of long genomes, even in pools containing relatively short VCG oligomers. For a genome of length LG, a sequence of at least nucleotides is required to uniquely specify a position within the genome. If the oligomers in the pool exceed this length by about three nucleotides, accurate replication becomes feasible (Fig. 2H). For example, genomes of length can be replicated in VCG pools containing 10 nt oligomers. However, LU equals only for a restricted set of genome sequences; more generally, . In such cases, reliable replication requires correspondingly longer oligomers. While the precise margin between oligomer length and LU depends on genome-specific features (particularly the motif distribution at sub-LU scales), it typically amounts to only a few additional nucleotides (Section S8 in the Supplementary Material).

It is noteworthy that replication in the VCG scenario imposes a selection pressure on prebiotic genomes to reduce their unique motif length, LU. A circular genome requiring many nucleotides to specify a unique locus (high LU) replicates less efficiently than one with a shorter LU, assuming all other properties of the VCG pool — particularly the oligomer length distribution — remain identical. This length distribution arises from the interplay between the chemical kinetics and molecular transport governed by the physical environment. For instance, templated ligation in an open system with continuous oligomer influx and outflux can produce a non-monotonic length distribution, with a concentration peak at a characteristic oligomer length Lc, determined by the interplay between dehybridization and outflux (Rosenberger et al., 2021). Through this emergent length scale, the environment shapes replication in the VCG scenario. If the environment facilitates long oligomers (Lc > LU), replication proceeds efficiently. Conversely, in environments with a small Lc, repeating motifs longer than Lc are selected against. In such cases, mutational errors may replace long repeated motifs with functionally equivalent sequences composed of shorter unique motifs, thereby increasing replication efficiency.

Given the broad range of prebiotically plausible non-equilibrium environments (Ianeselli et al., 2023), it is reasonable to expect that some environments provide the required conditions for efficient replication. The constraints formulated in this work can help to guide the search for self-replicating oligomer pools, in the vast space of possible concentration profiles and non-equilibrium environments.